---
title: "Alternative Splicing Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
params:
  comparison_id: "comparison"
  rmats_dir: NULL
  majiq_dir: NULL
  isar_dir: NULL
  sashimi_dir: NULL
  pegasas_dir: NULL
  leafcutter_dir: NULL
  fdr_cutoff: 0.05
  dpsi_cutoff: 0.1
  majiq_prob_threshold: 0.95
  majiq_dpsi_cutoff: 0.2
  nfcore_multiqc_dir: NULL
  organism: "human"
  de_results: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(knitr)
  library(DT)
  library(ggrepel)
  library(UpSetR)
})

# Initialize result containers; populated by tool-specific chunks below.
# Cross-tool and enrichment sections rely on these always existing.
all_rmats         <- data.frame()
majiq_results     <- data.frame()
isar_switches     <- data.frame()
isar_consequences <- data.frame()
leafcutter_clusters <- data.frame()
```

# Summary

This report presents differential alternative splicing analysis results for comparison: **`r params$comparison_id`**.

**Analysis tools:**

- rMATS-turbo: Differential alternative splicing (SE, A5SS, A3SS, MXE, RI events)
- MAJIQ V3: Local Splicing Variation (LSV) detection
- IsoformSwitchAnalyzeR: Isoform switch identification with functional consequences

**Filter criteria:**

- rMATS / ISAR FDR cutoff: `r params$fdr_cutoff`
- rMATS / ISAR |ΔΨ| / |dIF| cutoff: `r params$dpsi_cutoff`
- MAJIQ P(|ΔΨ| > threshold) cutoff: `r params$majiq_prob_threshold`
- MAJIQ |ΔΨ| cutoff: `r params$majiq_dpsi_cutoff`

---

# QC Metrics

```{r qc_load, eval=!is.null(params$nfcore_multiqc_dir)}
# Parse key QC metrics from nf-core/rnaseq MultiQC output.
# Expects the multiqc/star_salmon/ directory (contains multiqc_report_data/).
mqc_data_dir <- file.path(params$nfcore_multiqc_dir, "multiqc_report_data")

read_mqc_table <- function(path) {
  if (!file.exists(path)) return(data.frame())
  tryCatch(
    read.delim(path, stringsAsFactors = FALSE, check.names = FALSE),
    error = function(e) data.frame()
  )
}

qc_general <- read_mqc_table(file.path(mqc_data_dir, "multiqc_general_stats.txt"))
qc_star    <- read_mqc_table(file.path(mqc_data_dir, "multiqc_star.txt"))

# Remove per-read entries added by MultiQC (e.g. "Sample Read 1", "Sample Read 2")
if (nrow(qc_general) > 0) {
  qc_general <- qc_general[!grepl(" Read [12]$", qc_general$Sample), ]
}

# Build a clean per-sample QC table
qc_combined <- data.frame()

if (nrow(qc_star) > 0 && "Sample" %in% colnames(qc_star)) {
  qc_combined <- qc_star %>%
    select(Sample, any_of(c(
      "total_reads", "uniquely_mapped_percent"
    )))
}

if (nrow(qc_general) > 0 && "Sample" %in% colnames(qc_general)) {
  # Select available QC columns using flexible pattern matching
  wanted_patterns <- c(
    "PERCENT_DUPLICATION",
    "reads_aligned$",
    "reads_aligned_exonic",
    "reads_aligned_intronic",
    "5_3_bias"
  )
  gen_cols <- c("Sample", grep(
    paste(wanted_patterns, collapse = "|"),
    colnames(qc_general), value = TRUE
  ))
  gen_cols <- intersect(gen_cols, colnames(qc_general))

  if (length(gen_cols) > 1) {
    qc_sub <- qc_general[, gen_cols, drop = FALSE]
    # Shorten column names: strip tool prefix (e.g. "qualimap_rnaseq-reads_aligned" → "reads_aligned")
    colnames(qc_sub) <- sub("^[^-]+-", "", colnames(qc_sub))

    if (nrow(qc_combined) > 0) {
      qc_combined <- merge(qc_combined, qc_sub, by = "Sample", all = TRUE)
    } else {
      qc_combined <- qc_sub
    }
  }
}
```

```{r qc_table, eval=!is.null(params$nfcore_multiqc_dir) && nrow(qc_combined) > 0}
# Round numeric columns for display
qc_display <- qc_combined %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

datatable(
  qc_display,
  caption  = "Per-sample QC metrics from nf-core/rnaseq (MultiQC + STAR + Qualimap)",
  options  = list(pageLength = 20, scrollX = TRUE),
  rownames = FALSE
)
```

```{r qc_mapping_plot, eval=!is.null(params$nfcore_multiqc_dir) && nrow(qc_combined) > 0 && "uniquely_mapped_percent" %in% colnames(qc_combined), fig.height=5}
# Horizontal bar chart of uniquely mapped reads % per sample
qc_combined %>%
  filter(!is.na(uniquely_mapped_percent)) %>%
  mutate(Sample = reorder(Sample, uniquely_mapped_percent)) %>%
  ggplot(aes(x = Sample, y = uniquely_mapped_percent)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_hline(yintercept = 70, linetype = "dashed", color = "red", linewidth = 0.5) +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Uniquely mapped reads (%)",
    x     = NULL,
    y     = "Uniquely mapped (%)"
  )
```

```{r qc_dup_plot, eval=!is.null(params$nfcore_multiqc_dir) && nrow(qc_combined) > 0 && "PERCENT_DUPLICATION" %in% colnames(qc_combined), fig.height=5}
# Duplication rate per sample
qc_combined %>%
  filter(!is.na(PERCENT_DUPLICATION)) %>%
  mutate(
    Sample    = reorder(Sample, PERCENT_DUPLICATION),
    dup_pct   = PERCENT_DUPLICATION * 100
  ) %>%
  ggplot(aes(x = Sample, y = dup_pct)) +
  geom_col(fill = "coral", alpha = 0.8) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Duplication rate (%)",
    x     = NULL,
    y     = "Duplicate reads (%)"
  )
```

---

# rMATS Results

```{r rmats_load, eval=!is.null(params$rmats_dir)}
event_types <- c("SE", "A5SS", "A3SS", "MXE", "RI")
rmats_results_list <- list()

for (event_type in event_types) {
  jc_file <- file.path(params$rmats_dir, paste0(event_type, ".MATS.JC.txt"))
  if (file.exists(jc_file)) {
    df <- read.delim(jc_file, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
      df$EventType <- event_type
      rmats_results_list[[event_type]] <- df
    }
  }
}

if (length(rmats_results_list) > 0) {
  all_rmats <- bind_rows(rmats_results_list) %>%
    mutate(
      Significant    = FDR <= params$fdr_cutoff & abs(IncLevelDifference) >= params$dpsi_cutoff,
      # Priority score: combines -log10(FDR) with effect size.
      # Higher = more significant and larger PSI change.
      priority_score = round(-log10(pmax(FDR, 1e-300)) * abs(IncLevelDifference), 3)
    )
}
```

```{r rmats_summary, eval=!is.null(params$rmats_dir) && nrow(all_rmats) > 0}
summary_table <- all_rmats %>%
  group_by(EventType) %>%
  summarise(
    Total       = n(),
    Significant = sum(Significant, na.rm = TRUE),
    FDR_leq_0.05 = sum(FDR <= 0.05, na.rm = TRUE),
    Mean_deltaPSI = round(mean(abs(IncLevelDifference), na.rm = TRUE), 3)
  )

kable(summary_table, caption = "rMATS Event Summary by Event Type")
```

```{r rmats_volcano, eval=!is.null(params$rmats_dir) && nrow(all_rmats) > 0, fig.height=8}
# Label top 5 events per event type (highest priority score)
top_labels <- all_rmats %>%
  filter(Significant, !is.na(geneSymbol)) %>%
  group_by(EventType) %>%
  slice_max(priority_score, n = 5) %>%
  ungroup()

all_rmats %>%
  filter(!is.na(FDR) & FDR > 0) %>%
  ggplot(aes(x = IncLevelDifference, y = -log10(FDR), color = Significant)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_text_repel(
    data = top_labels,
    aes(label = geneSymbol),
    size = 2.5, max.overlaps = 10, color = "black"
  ) +
  facet_wrap(~EventType, scales = "free_y") +
  scale_color_manual(values = c("gray70", "firebrick")) +
  geom_hline(yintercept = -log10(params$fdr_cutoff),    linetype = "dashed", color = "navy", linewidth = 0.4) +
  geom_vline(xintercept = c(-params$dpsi_cutoff, params$dpsi_cutoff), linetype = "dashed", color = "navy", linewidth = 0.4) +
  theme_bw() +
  labs(
    title = "rMATS Volcano Plots",
    x     = "ΔΨ (IncLevelDifference)",
    y     = "-log10(FDR)"
  ) +
  theme(legend.position = "bottom")
```

```{r rmats_top_table, eval=!is.null(params$rmats_dir) && nrow(all_rmats) > 0}
top_events <- all_rmats %>%
  filter(Significant) %>%
  arrange(desc(priority_score)) %>%
  select(EventType, GeneID, geneSymbol, chr, strand, FDR, IncLevelDifference, priority_score) %>%
  head(50)

if (nrow(top_events) > 0) {
  datatable(
    top_events,
    caption  = "Top 50 significant rMATS events ranked by priority score",
    options  = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    formatRound(columns = c("FDR", "IncLevelDifference", "priority_score"), digits = 4)
} else {
  cat("No significant events detected.\n")
}
```

---

# MAJIQ Results

```{r majiq_load, eval=!is.null(params$majiq_dir)}
tsv_files <- list.files(params$majiq_dir, pattern = "\\.tsv$", full.names = TRUE)
tsv_file  <- if (length(tsv_files) > 0) tsv_files[1] else NA_character_

if (!is.na(tsv_file) && file.exists(tsv_file)) {
  majiq_results <- tryCatch(
    # comment.char skips the JSON metadata block (#-prefixed lines) at the top of MAJIQ v3 TSVs
    read.delim(tsv_file, stringsAsFactors = FALSE, comment.char = "#"),
    error = function(e) data.frame()
  )

  if (nrow(majiq_results) > 0 &&
      "probability_changing" %in% colnames(majiq_results) &&
      "dpsi_mean" %in% colnames(majiq_results)) {
    majiq_results <- majiq_results %>%
      mutate(Significant = probability_changing >= params$majiq_prob_threshold &
               abs(dpsi_mean) >= params$majiq_dpsi_cutoff)
  } else {
    majiq_results$Significant <- FALSE
  }
}
```

```{r majiq_summary, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0}
cat("Total junctions detected:", nrow(majiq_results), "\n\n")
if ("Significant" %in% colnames(majiq_results)) {
  cat(sprintf(
    "Significant LSVs (P(|ΔPSI|) >= %.2f & |ΔPSI| >= %.2f): %d\n\n",
    params$majiq_prob_threshold,
    params$majiq_dpsi_cutoff,
    sum(majiq_results$Significant, na.rm = TRUE)
  ))
}
```

```{r majiq_distribution, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0 && "dpsi_mean" %in% colnames(majiq_results)}
majiq_results %>%
  filter(!is.na(dpsi_mean)) %>%
  ggplot(aes(x = abs(dpsi_mean))) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  theme_bw() +
  labs(
    title = "Distribution of |ΔΨ| in MAJIQ junctions",
    x     = "|dpsi_mean|",
    y     = "Count"
  )
```

```{r majiq_top_table, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0}
cols_to_show <- intersect(
  c("gene_name", "gene_id", "event_type", "seqid", "start", "end",
    "dpsi_mean", "dpsi_std", "probability_changing", "probability_nonchanging",
    "grp1_raw_psi_mean", "grp2_raw_psi_mean"),
  colnames(majiq_results)
)

if (length(cols_to_show) > 0) {
  top_lsvs <- majiq_results %>%
    filter(Significant) %>%
    arrange(desc(probability_changing), desc(abs(dpsi_mean))) %>%
    select(all_of(cols_to_show)) %>%
    head(100)

  if (nrow(top_lsvs) > 0) {
    datatable(
      top_lsvs,
      caption  = paste0("Top 100 significant MAJIQ junctions (P >= ",
                        params$majiq_prob_threshold, ", |ΔΨ| >= ",
                        params$majiq_dpsi_cutoff, ")"),
      options  = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      formatRound(columns = names(top_lsvs)[sapply(top_lsvs, is.numeric)], digits = 4)
  } else {
    cat("No significant junctions detected.\n")
  }
}
```

---

# IsoformSwitchAnalyzeR Results

```{r isar_load, eval=!is.null(params$isar_dir)}
read_csv_safe <- function(path) {
  if (!file.exists(path) || file.size(path) < 3L) return(data.frame())
  tryCatch(read.csv(path, stringsAsFactors = FALSE), error = function(e) data.frame())
}

isar_switches     <- read_csv_safe(file.path(params$isar_dir, "top_isoform_switches.csv"))
isar_consequences <- read_csv_safe(file.path(params$isar_dir, "consequence_summary.csv"))
```

```{r isar_summary, eval=!is.null(params$isar_dir) && nrow(isar_switches) > 0}
if ("isoform_switch_q_value" %in% colnames(isar_switches) &&
    "dIF" %in% colnames(isar_switches)) {
  isar_switches$significant <- (
    isar_switches$isoform_switch_q_value < params$fdr_cutoff &
    abs(isar_switches$dIF) >= params$dpsi_cutoff
  )
} else {
  isar_switches$significant <- NA
}

n_total <- nrow(isar_switches)
n_sig   <- sum(isar_switches$significant, na.rm = TRUE)

cat("Total isoforms tested:", n_total, "\n\n")
cat(sprintf("Significant (FDR < %.2f & |dIF| >= %.2f): %d\n\n",
            params$fdr_cutoff, params$dpsi_cutoff, n_sig))

if ("gene_name" %in% colnames(isar_switches)) {
  cat("Genes tested:", length(unique(isar_switches$gene_name)), "\n")
  if (n_sig > 0) {
    sig_genes_isar_n <- length(unique(isar_switches$gene_name[isar_switches$significant]))
    cat("Genes with significant switches:", sig_genes_isar_n, "\n")
  }
}
```

```{r isar_consequences_plot, eval=!is.null(params$isar_dir) && nrow(isar_consequences) > 0, fig.height=6}
if ("feature" %in% colnames(isar_consequences) && "count" %in% colnames(isar_consequences)) {
  isar_consequences %>%
    ggplot(aes(x = reorder(feature, count), y = count)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    theme_bw() +
    labs(
      title = "Functional Consequences of Isoform Switches",
      x     = "Consequence Type",
      y     = "Count"
    )
}
```

```{r isar_top_table, eval=!is.null(params$isar_dir) && nrow(isar_switches) > 0}
cols_to_show <- intersect(
  c("gene_name", "gene_id", "isoform_id", "condition_1", "condition_2",
    "IF1", "IF2", "dIF", "isoform_switch_q_value", "gene_switch_q_value", "significant"),
  colnames(isar_switches)
)

if (length(cols_to_show) > 0) {
  display_df <- isar_switches %>% select(all_of(cols_to_show))
  has_sig     <- "significant" %in% colnames(display_df)
  n_sig_table <- if (has_sig) sum(display_df$significant, na.rm = TRUE) else 0

  dt <- datatable(
    display_df,
    caption  = paste0(
      "All tested isoforms, ordered by q-value. ",
      if (has_sig) paste0(n_sig_table, " pass FDR < ", params$fdr_cutoff,
                          " & |dIF| >= ", params$dpsi_cutoff, ".") else ""
    ),
    options  = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    formatRound(columns = names(display_df)[sapply(display_df, is.numeric)], digits = 4)

  if (has_sig) {
    dt <- dt %>%
      formatStyle(
        "significant",
        target          = "row",
        backgroundColor = styleEqual(c(TRUE, FALSE), c("#d4edda", "white"))
      )
  }
  dt
}
```

---

# LeafCutter Results

```{r leafcutter_load, eval=!is.null(params$leafcutter_dir)}
# Standard leafcutter_ds.R output: *_cluster_significance.txt and *_effect_sizes.txt
sig_file <- list.files(params$leafcutter_dir,
                       pattern = "_cluster_significance\\.txt$",
                       full.names = TRUE)
sig_file <- if (length(sig_file) > 0) sig_file[1] else NA_character_

eff_file <- list.files(params$leafcutter_dir,
                       pattern = "_effect_sizes\\.txt$",
                       full.names = TRUE)
eff_file <- if (length(eff_file) > 0) eff_file[1] else NA_character_

if (!is.na(sig_file) && file.exists(sig_file)) {
  leafcutter_clusters <- tryCatch(
    read.delim(sig_file, stringsAsFactors = FALSE),
    error = function(e) data.frame()
  )
  if (nrow(leafcutter_clusters) > 0 && "p.adjust" %in% colnames(leafcutter_clusters)) {
    leafcutter_clusters <- leafcutter_clusters %>%
      mutate(Significant = p.adjust < params$fdr_cutoff & status == "OK")
  }
}

leafcutter_effects <- data.frame()
if (!is.na(eff_file) && file.exists(eff_file)) {
  leafcutter_effects <- tryCatch(
    read.delim(eff_file, stringsAsFactors = FALSE),
    error = function(e) data.frame()
  )
}
```

```{r leafcutter_summary, eval=!is.null(params$leafcutter_dir) && nrow(leafcutter_clusters) > 0}
n_clusters <- nrow(leafcutter_clusters)
n_ok       <- sum(leafcutter_clusters$status == "OK", na.rm = TRUE)
n_sig      <- sum(leafcutter_clusters$Significant, na.rm = TRUE)

cat(sprintf("Total clusters tested: %d  \n", n_clusters))
cat(sprintf("Clusters passing noise filter (status = OK): %d  \n", n_ok))
cat(sprintf("Significant clusters (FDR < %.2f): %d\n\n", params$fdr_cutoff, n_sig))

if ("genes" %in% colnames(leafcutter_clusters)) {
  sig_genes_lc <- unique(unlist(strsplit(
    leafcutter_clusters$genes[leafcutter_clusters$Significant & !is.na(leafcutter_clusters$genes)],
    ","
  )))
  sig_genes_lc <- trimws(sig_genes_lc[sig_genes_lc != "" & !is.na(sig_genes_lc)])
  cat(sprintf("Genes in significant clusters: %d\n", length(sig_genes_lc)))
}
```

```{r leafcutter_dpsi_dist, eval=!is.null(params$leafcutter_dir) && nrow(leafcutter_effects) > 0 && "deltapsi" %in% colnames(leafcutter_effects), fig.height=5}
leafcutter_effects %>%
  filter(!is.na(deltapsi)) %>%
  ggplot(aes(x = abs(deltapsi))) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  theme_bw() +
  labs(
    title = "Distribution of |ΔΨ| across LeafCutter introns",
    x     = "|ΔΨ|",
    y     = "Count"
  )
```

```{r leafcutter_top_table, eval=!is.null(params$leafcutter_dir) && nrow(leafcutter_clusters) > 0}
cols_to_show <- intersect(
  c("cluster", "genes", "status", "loglr", "df", "p", "p.adjust", "Significant"),
  colnames(leafcutter_clusters)
)

if (length(cols_to_show) > 0) {
  top_clusters <- leafcutter_clusters %>%
    filter(status == "OK") %>%
    arrange(p.adjust) %>%
    select(all_of(cols_to_show)) %>%
    head(100)

  if (nrow(top_clusters) > 0) {
    datatable(
      top_clusters,
      caption  = paste0("Top 100 LeafCutter clusters (status = OK), sorted by FDR. ",
                        n_sig, " pass FDR < ", params$fdr_cutoff, "."),
      options  = list(pageLength = 15, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      formatRound(columns = intersect(c("loglr", "p", "p.adjust"), cols_to_show), digits = 4)
  } else {
    cat("No clusters passed the noise filter.\n")
  }
}
```

---

# Cross-Tool Overlap

```{r overlap_build}
# Build per-tool gene sets of significant AS genes
gene_sets <- list()

rmats_sig_genes <- character(0)
if (nrow(all_rmats) > 0 && "geneSymbol" %in% colnames(all_rmats)) {
  rmats_sig_genes <- unique(all_rmats$geneSymbol[
    all_rmats$Significant & !is.na(all_rmats$geneSymbol) & all_rmats$geneSymbol != ""
  ])
  if (length(rmats_sig_genes) > 0) gene_sets[["rMATS"]] <- rmats_sig_genes
}

majiq_sig_genes <- character(0)
if (nrow(majiq_results) > 0 && "gene_name" %in% colnames(majiq_results)) {
  majiq_sig_genes <- unique(majiq_results$gene_name[
    majiq_results$Significant & !is.na(majiq_results$gene_name) & majiq_results$gene_name != ""
  ])
  if (length(majiq_sig_genes) > 0) gene_sets[["MAJIQ"]] <- majiq_sig_genes
}

isar_sig_genes <- character(0)
if (nrow(isar_switches) > 0 && "gene_name" %in% colnames(isar_switches) &&
    "significant" %in% colnames(isar_switches)) {
  isar_sig_genes <- unique(isar_switches$gene_name[
    isar_switches$significant & !is.na(isar_switches$gene_name)
  ])
  if (length(isar_sig_genes) > 0) gene_sets[["ISAR"]] <- isar_sig_genes
}

leafcutter_sig_genes <- character(0)
if (nrow(leafcutter_clusters) > 0 &&
    "genes" %in% colnames(leafcutter_clusters) &&
    "Significant" %in% colnames(leafcutter_clusters)) {
  lc_genes_raw <- leafcutter_clusters$genes[
    leafcutter_clusters$Significant & !is.na(leafcutter_clusters$genes)
  ]
  leafcutter_sig_genes <- unique(trimws(unlist(strsplit(lc_genes_raw, ","))))
  leafcutter_sig_genes <- leafcutter_sig_genes[leafcutter_sig_genes != ""]
  if (length(leafcutter_sig_genes) > 0) gene_sets[["LeafCutter"]] <- leafcutter_sig_genes
}
```

```{r overlap_upset, eval=length(gene_sets) >= 2, fig.height=5}
upset(
  fromList(gene_sets),
  order.by    = "freq",
  sets.bar.color = "steelblue",
  main.bar.color = "steelblue"
)
```

```{r overlap_summary, eval=length(gene_sets) >= 2}
# Genes found significant by all available tools
all_tools_genes <- Reduce(intersect, gene_sets)
cat(sprintf("Genes significant in all %d active tools: %d\n\n",
            length(gene_sets), length(all_tools_genes)))

if (length(all_tools_genes) > 0) {
  cat("Genes:", paste(sort(all_tools_genes), collapse = ", "), "\n")
}
```

```{r overlap_pairwise, eval=length(gene_sets) >= 2}
# Pairwise overlap table
tool_names <- names(gene_sets)
overlap_mat <- matrix(
  nrow = length(tool_names), ncol = length(tool_names),
  dimnames = list(tool_names, tool_names)
)
for (i in tool_names) {
  for (j in tool_names) {
    overlap_mat[i, j] <- length(intersect(gene_sets[[i]], gene_sets[[j]]))
  }
}
kable(as.data.frame(overlap_mat),
      caption = "Pairwise gene overlap between tools (number of shared significant genes)")
```

---

# PSI Profile Analysis

## PCA of splicing profiles

```{r psi_pca, eval=nrow(all_rmats) > 0, fig.height=6}
# Build per-sample PSI matrix from rMATS IncLevel columns (SE events preferred; fallback to all).
# IncLevel1/IncLevel2 are comma-separated PSI values per replicate in each group.

parse_psi_vec <- function(x) {
  if (is.na(x) || x == "NA" || x == "") return(NA_real_)
  suppressWarnings(as.numeric(strsplit(as.character(x), ",")[[1]]))
}

# Use SE events if available; otherwise fall back to whichever event type has most significant events
pca_event_type <- if ("SE" %in% all_rmats$EventType[all_rmats$Significant]) "SE" else {
  all_rmats %>% filter(Significant) %>% count(EventType) %>% slice_max(n, n=1) %>% pull(EventType)
}
pca_events_raw <- if (length(pca_event_type) > 0) {
  all_rmats %>%
    filter(Significant, EventType == pca_event_type[1],
           !is.na(IncLevel1), !is.na(IncLevel2),
           IncLevel1 != "NA", IncLevel2 != "NA") %>%
    head(2000)   # cap for performance
} else data.frame()

if (nrow(pca_events_raw) >= 10) {
  psi_list <- lapply(seq_len(nrow(pca_events_raw)), function(i) {
    c(parse_psi_vec(pca_events_raw$IncLevel1[i]),
      parse_psi_vec(pca_events_raw$IncLevel2[i]))
  })

  # Only keep events where every sample has a valid PSI
  n_samples <- length(psi_list[[1]])
  valid_events <- sapply(psi_list, function(v) length(v) == n_samples && !any(is.na(v)))
  psi_list <- psi_list[valid_events]

  if (length(psi_list) >= 10) {
    psi_mat <- do.call(rbind, psi_list)   # events × samples

    # Determine group sizes from the first valid event
    n1 <- length(parse_psi_vec(pca_events_raw$IncLevel1[which(valid_events)[1]]))
    n2 <- length(parse_psi_vec(pca_events_raw$IncLevel2[which(valid_events)[1]]))

    sample_labels <- c(paste0("Group1_rep", seq_len(n1)), paste0("Group2_rep", seq_len(n2)))
    group_labels  <- c(rep("Group1", n1), rep("Group2", n2))

    # PCA on transposed matrix (samples as rows)
    pca_res <- tryCatch(
      prcomp(t(psi_mat), center = TRUE, scale. = FALSE),
      error = function(e) NULL
    )

    if (!is.null(pca_res)) {
      var_exp  <- round(summary(pca_res)$importance[2, 1:min(2, ncol(pca_res$x))] * 100, 1)
      pca_df   <- as.data.frame(pca_res$x[, 1:min(2, ncol(pca_res$x))])
      pca_df$Group  <- group_labels
      pca_df$Sample <- sample_labels

      ggplot(pca_df, aes(x = PC1, y = PC2, color = Group, label = Sample)) +
        geom_point(size = 3) +
        geom_text_repel(size = 3, max.overlaps = 20) +
        theme_bw() +
        scale_color_brewer(palette = "Set1") +
        labs(
          title    = paste0("PCA of PSI profiles (", pca_event_type[1], " events, n=", length(psi_list), ")"),
          x        = paste0("PC1 (", var_exp[1], "% variance)"),
          y        = paste0("PC2 (", var_exp[2], "% variance)")
        )
    }
  } else {
    cat("Insufficient events with complete PSI data for PCA (need >= 10).\n")
  }
} else {
  cat("No significant rMATS events available for PCA.\n")
}
```

## Splice junction coverage QC

```{r junction_qc, eval=nrow(all_rmats) > 0, fig.height=5}
# Per-sample mean inclusion junction count (IJC) as a proxy for splice junction coverage.
# Low values flag samples with insufficient coverage for reliable AS detection.

parse_sample_counts <- function(count_str, group_label) {
  if (is.na(count_str) || count_str == "") return(NULL)
  vals <- suppressWarnings(as.integer(strsplit(as.character(count_str), ",")[[1]]))
  if (length(vals) == 0) return(NULL)
  data.frame(
    sample_idx = seq_along(vals),
    count      = vals,
    group      = group_label,
    stringsAsFactors = FALSE
  )
}

jq_events <- all_rmats %>%
  filter(EventType == "SE", !is.na(IJC_SAMPLE_1), !is.na(IJC_SAMPLE_2)) %>%
  head(500)

if (nrow(jq_events) > 0) {
  jq_raw <- bind_rows(lapply(seq_len(nrow(jq_events)), function(i) {
    bind_rows(
      parse_sample_counts(jq_events$IJC_SAMPLE_1[i], "Group1"),
      parse_sample_counts(jq_events$IJC_SAMPLE_2[i], "Group2")
    )
  }))

  if (nrow(jq_raw) > 0) {
    jq_summary <- jq_raw %>%
      group_by(group, sample_idx) %>%
      summarise(mean_IJC = mean(count, na.rm = TRUE), .groups = "drop") %>%
      mutate(Sample = paste0(group, "_rep", sample_idx))

    ggplot(jq_summary, aes(x = reorder(Sample, mean_IJC), y = mean_IJC, fill = group)) +
      geom_col(alpha = 0.8) +
      coord_flip() +
      theme_bw() +
      scale_fill_brewer(palette = "Set1") +
      labs(
        title = "Mean inclusion junction count per sample (SE events, top 500)",
        subtitle = "Low values may indicate insufficient coverage for reliable AS detection",
        x = NULL,
        y = "Mean IJC"
      )
  }
} else {
  cat("No SE events with junction count data available.\n")
}
```

---

# Pathway Enrichment

```{r enrichment_setup}
# Pool significant genes from all active tools
all_sig_genes <- unique(c(rmats_sig_genes, isar_sig_genes, leafcutter_sig_genes))
has_genes_for_enrichment <- length(all_sig_genes) >= 10
```

```{r go_enrichment, eval=has_genes_for_enrichment}
suppressPackageStartupMessages(library(clusterProfiler))

# Select annotation database based on organism param
orgdb_name <- switch(
  params$organism,
  "mouse" = "org.Mm.eg.db",
  "org.Hs.eg.db"   # default: human
)
suppressPackageStartupMessages(library(orgdb_name, character.only = TRUE))

ego <- tryCatch(
  enrichGO(
    gene          = all_sig_genes,
    OrgDb         = get(orgdb_name),
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2,
    readable      = TRUE
  ),
  error = function(e) {
    message("GO enrichment failed: ", conditionMessage(e))
    NULL
  }
)
```

```{r go_dotplot, eval=has_genes_for_enrichment && !is.null(ego) && nrow(ego) > 0, fig.height=8}
dotplot(ego, showCategory = 25) +
  labs(title = paste0("GO Biological Process enrichment (n=", length(all_sig_genes), " AS genes)")) +
  theme(axis.text.y = element_text(size = 9))
```

```{r go_table, eval=has_genes_for_enrichment && !is.null(ego) && nrow(ego) > 0}
ego_df <- as.data.frame(ego) %>%
  select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
  mutate(across(c(pvalue, p.adjust), ~ signif(.x, 3)))

datatable(
  ego_df,
  caption  = "GO Biological Process enrichment (p.adjust < 0.05)",
  options  = list(pageLength = 15, scrollX = TRUE),
  rownames = FALSE
)
```

```{r kegg_enrichment, eval=has_genes_for_enrichment}
# KEGG enrichment — requires gene symbol → Entrez ID mapping
gene_ids <- tryCatch(
  bitr(all_sig_genes, fromType = "SYMBOL", toType = "ENTREZID",
       OrgDb = get(orgdb_name)),
  error = function(e) data.frame()
)

ekegg <- NULL
if (nrow(gene_ids) >= 5) {
  kegg_organism <- switch(params$organism, "mouse" = "mmu", "hsa")
  ekegg <- tryCatch(
    enrichKEGG(
      gene          = gene_ids$ENTREZID,
      organism      = kegg_organism,
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05
    ),
    error = function(e) NULL
  )
}
```

```{r kegg_dotplot, eval=has_genes_for_enrichment && !is.null(ekegg) && nrow(ekegg) > 0, fig.height=6}
dotplot(ekegg, showCategory = 20) +
  labs(title = "KEGG pathway enrichment (AS genes)") +
  theme(axis.text.y = element_text(size = 9))
```

```{r no_enrichment_msg, eval=!has_genes_for_enrichment}
cat("Pathway enrichment requires at least 10 significant AS genes across tools.",
    "Increase the number of significant events or lower the FDR / |ΔΨ| cutoffs.\n")
```

---

# Sashimi Plots

```{r sashimi_section, eval=!is.null(params$sashimi_dir)}
# List available sashimi plot PDFs
sashimi_pdfs <- list.files(params$sashimi_dir, pattern = "\\.pdf$",
                           recursive = TRUE, full.names = TRUE)

if (length(sashimi_pdfs) > 0) {
  cat(sprintf("Sashimi plots generated for %d events.\n\n", length(sashimi_pdfs)))
  cat("PDFs are available in the sashimi_plots/ output directory.\n")

  # Build a summary table of plotted events
  sashimi_table <- data.frame(
    file = basename(sashimi_pdfs),
    path = sashimi_pdfs,
    stringsAsFactors = FALSE
  )
  datatable(sashimi_table, caption = "Sashimi plot files",
            options = list(pageLength = 20), rownames = FALSE)
} else {
  cat("No sashimi plot PDFs found in the output directory.\n")
}
```

---

# PEGASAS: Pathway–Splicing Correlation

```{r pegasas_check}
run_pegasas_section <- !is.null(params$pegasas_dir)
```

```{r pegasas_scores, eval=run_pegasas_section}
# Load per-sample KS enrichment scores produced by PEGASAS pathway step.
# Each subdirectory under pegasas_dir/correlation_out/ corresponds to one gene signature.
corr_dirs <- list.dirs(file.path(params$pegasas_dir), recursive = FALSE, full.names = TRUE)

pegasas_score_plots <- list()
pegasas_cor_tables  <- list()

for (cdir in corr_dirs) {
  sig_name <- basename(cdir)

  # Pathway scores file: <sig>.sorted.txt (sample × score, 2-row: header + values)
  scores_file <- file.path(cdir, paste0(sig_name, ".sorted.txt"))
  if (!file.exists(scores_file)) next

  scores_raw <- read.table(scores_file, sep = "\t", header = FALSE,
                           stringsAsFactors = FALSE)
  # Row 1 = sample IDs; Row 2 = KS scores
  if (nrow(scores_raw) < 2) next
  samples    <- unlist(scores_raw[1, -1])
  ks_scores  <- as.numeric(unlist(scores_raw[2, -1]))
  score_df   <- data.frame(sample = samples, ks_score = ks_scores,
                            stringsAsFactors = FALSE)

  p <- ggplot(score_df, aes(x = reorder(sample, ks_score), y = ks_score)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(title = sig_name,
         x = NULL,
         y = "KS enrichment score") +
    theme_bw(base_size = 11)
  pegasas_score_plots[[sig_name]] <- p

  # High-correlation matrix
  high_cor_file <- file.path(cdir, paste0(sig_name, "_high_cor_matrix.txt"))
  if (file.exists(high_cor_file)) {
    cor_df <- tryCatch(
      read.table(high_cor_file, sep = "\t", header = TRUE,
                 stringsAsFactors = FALSE),
      error = function(e) NULL
    )
    if (!is.null(cor_df) && nrow(cor_df) > 0) {
      pegasas_cor_tables[[sig_name]] <- cor_df
    }
  }
}

if (length(pegasas_score_plots) == 0) {
  cat("No PEGASAS pathway score files found.\n")
}
```

```{r pegasas_score_plots, eval=run_pegasas_section && length(pegasas_score_plots) > 0, results='asis'}
cat("## Pathway KS Enrichment Scores\n\n")
cat("Per-sample KS enrichment scores (higher = stronger pathway activity).\n\n")
for (sig in names(pegasas_score_plots)) {
  cat(sprintf("### %s\n\n", sig))
  print(pegasas_score_plots[[sig]])
  cat("\n\n")
}
```

```{r pegasas_cor_tables, eval=run_pegasas_section && length(pegasas_cor_tables) > 0, results='asis'}
cat("## Pathway-Correlated Splicing Events\n\n")
cat("Events with |Pearson r| > 0.3 and permutation p < 0.0002.\n\n")
for (sig in names(pegasas_cor_tables)) {
  cat(sprintf("### %s\n\n", sig))
  print(
    datatable(
      pegasas_cor_tables[[sig]],
      caption   = paste("High-correlation splicing events:", sig),
      rownames  = FALSE,
      filter    = "top",
      options   = list(pageLength = 15, scrollX = TRUE)
    )
  )
  cat("\n\n")
}
```

---

# DE + AS Integration

```{r de_load}
# Load differential expression results.
# Expected format: TSV with columns gene_id, log2FC, padj.
# gene_id should match gene symbols used by rMATS (geneSymbol), MAJIQ (gene_name), and ISAR.
run_de_section <- !is.null(params$de_results) && file.exists(params$de_results)
de_results     <- data.frame()
de_sig_genes   <- character(0)

if (run_de_section) {
  de_results <- tryCatch(
    read.delim(params$de_results, stringsAsFactors = FALSE),
    error = function(e) {
      message("Could not load DE results: ", conditionMessage(e))
      data.frame()
    }
  )

  # Validate required columns
  required_cols <- c("gene_id", "log2FC", "padj")
  missing_cols  <- setdiff(required_cols, colnames(de_results))
  if (length(missing_cols) > 0) {
    warning("DE results file is missing columns: ", paste(missing_cols, collapse = ", "),
            ". DE+AS section will be skipped.")
    run_de_section <- FALSE
    de_results     <- data.frame()
  }
}

if (run_de_section && nrow(de_results) > 0) {
  de_results <- de_results %>%
    filter(!is.na(padj) & !is.na(log2FC)) %>%
    mutate(
      DE_significant = padj < params$fdr_cutoff & abs(log2FC) >= 1,
      neg_log10_padj = -log10(pmax(padj, 1e-300))
    )
  de_sig_genes <- unique(de_results$gene_id[de_results$DE_significant])
}
```

```{r de_as_overlap, eval=run_de_section && nrow(de_results) > 0}
# Pool AS significant genes from all active tools.
# Reuse the gene sets built in the Cross-Tool Overlap section.
as_sig_genes <- unique(c(rmats_sig_genes, majiq_sig_genes, isar_sig_genes, leafcutter_sig_genes))

# Tag each gene in the DE table by its AS status
de_results <- de_results %>%
  mutate(
    AS_hit = gene_id %in% as_sig_genes,
    hit_class = case_when(
      DE_significant & AS_hit  ~ "DE + AS",
      DE_significant & !AS_hit ~ "DE only",
      !DE_significant & AS_hit ~ "AS only",
      TRUE                     ~ "Neither"
    )
  )

dual_hits <- de_results %>%
  filter(hit_class == "DE + AS") %>%
  arrange(padj)

cat(sprintf(
  "DE significant (|log2FC| ≥ 1, padj < %.2f): **%d genes**  \n",
  params$fdr_cutoff, length(de_sig_genes)
))
cat(sprintf(
  "AS significant (any tool, FDR/prob cutoff): **%d genes**  \n",
  length(as_sig_genes)
))
cat(sprintf(
  "Dual hits (DE + AS): **%d genes**\n\n",
  nrow(dual_hits)
))
```

```{r de_volcano, eval=run_de_section && nrow(de_results) > 0, fig.height=7, fig.width=10}
# Volcano plot with AS-hit genes highlighted.
# Points are down-sampled for "Neither" to keep the plot readable.
set.seed(42)
de_plot <- bind_rows(
  de_results %>% filter(hit_class != "Neither"),
  de_results %>% filter(hit_class == "Neither") %>% slice_sample(n = min(3000, sum(de_results$hit_class == "Neither")))
)

label_df <- de_results %>%
  filter(hit_class == "DE + AS") %>%
  slice_max(neg_log10_padj, n = 20)

colour_map <- c(
  "DE + AS"  = "firebrick",
  "DE only"  = "steelblue",
  "AS only"  = "darkorange",
  "Neither"  = "gray80"
)
size_map   <- c("DE + AS" = 2.5, "DE only" = 1.5, "AS only" = 2, "Neither" = 0.6)
alpha_map  <- c("DE + AS" = 0.9, "DE only" = 0.6, "AS only" = 0.7, "Neither" = 0.3)

ggplot(de_plot, aes(x = log2FC, y = neg_log10_padj,
                    colour = hit_class, size = hit_class, alpha = hit_class)) +
  geom_point() +
  geom_text_repel(
    data        = label_df,
    aes(label   = gene_id),
    colour      = "black",
    size        = 2.8,
    max.overlaps = 15,
    segment.size = 0.3
  ) +
  scale_colour_manual(values = colour_map, name = NULL) +
  scale_size_manual(values = size_map,     name = NULL) +
  scale_alpha_manual(values = alpha_map,   name = NULL) +
  geom_hline(yintercept = -log10(params$fdr_cutoff),
             linetype = "dashed", colour = "gray40", linewidth = 0.4) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", colour = "gray40", linewidth = 0.4) +
  theme_bw(base_size = 12) +
  labs(
    title    = "Differential Expression — AS genes highlighted",
    subtitle = sprintf("Red = significant in both DE and AS (n = %d)", nrow(dual_hits)),
    x        = "log2 Fold Change",
    y        = "-log10(adj. p-value)"
  ) +
  theme(legend.position = "bottom")
```

```{r de_as_table, eval=run_de_section && nrow(dual_hits) > 0}
# Enrich dual-hit table with AS event details from rMATS (if available)
if (nrow(all_rmats) > 0 && "geneSymbol" %in% colnames(all_rmats)) {
  # Summarise per gene: best event type, lowest FDR, largest |dPSI|
  rmats_gene_summary <- all_rmats %>%
    filter(Significant, geneSymbol %in% dual_hits$gene_id) %>%
    group_by(geneSymbol) %>%
    slice_min(FDR, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(geneSymbol, EventType, FDR_AS = FDR,
           dPSI = IncLevelDifference, priority_score)

  dual_hits_ann <- dual_hits %>%
    left_join(rmats_gene_summary, by = c("gene_id" = "geneSymbol"))
} else {
  dual_hits_ann <- dual_hits
}

show_cols <- intersect(
  c("gene_id", "log2FC", "padj", "EventType", "FDR_AS", "dPSI", "priority_score"),
  colnames(dual_hits_ann)
)

datatable(
  dual_hits_ann[, show_cols],
  caption  = sprintf("Genes significant in both DE and AS (n = %d)", nrow(dual_hits_ann)),
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, scrollX = TRUE)
) %>%
  formatRound(columns = intersect(c("log2FC", "padj", "FDR_AS", "dPSI", "priority_score"),
                                   show_cols),
              digits = 4)
```

```{r de_as_barplot, eval=run_de_section && nrow(dual_hits) > 0 && nrow(all_rmats) > 0, fig.height=5}
# Distribution of AS event types among dual-hit genes
if ("EventType" %in% colnames(dual_hits_ann) && any(!is.na(dual_hits_ann$EventType))) {
  dual_hits_ann %>%
    filter(!is.na(EventType)) %>%
    count(EventType) %>%
    ggplot(aes(x = reorder(EventType, n), y = n, fill = EventType)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    theme_bw(base_size = 12) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "AS event types in DE+AS dual-hit genes",
      x     = NULL,
      y     = "Number of genes"
    )
}
```

```{r de_as_skip_msg, eval=!run_de_section}
cat("DE+AS integration requires a `--de_results` TSV file with columns: gene_id, log2FC, padj.  \n")
cat("Pass the path via `--de_results` to enable this section.\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
