---
title: "Alternative Splicing Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
params:
  comparison_id: "comparison"
  rmats_dir: NULL
  majiq_dir: NULL
  isar_dir: NULL
  fdr_cutoff: 0.05
  dpsi_cutoff: 0.1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(DT)
```

# Summary

This report presents differential alternative splicing analysis results for comparison: **`r params$comparison_id`**.

**Analysis tools:**

- rMATS-turbo: Differential alternative splicing (SE, A5SS, A3SS, MXE, RI events)
- MAJIQ V3: Local Splicing Variation (LSV) detection
- IsoformSwitchAnalyzeR: Isoform switch identification with functional consequences

**Filter criteria:**

- FDR cutoff: `r params$fdr_cutoff`
- |ΔΨ| cutoff: `r params$dpsi_cutoff`

---

# rMATS Results

```{r rmats_load, eval=!is.null(params$rmats_dir)}
# Load rMATS summary
event_types <- c("SE", "A5SS", "A3SS", "MXE", "RI")
rmats_results <- list()

for (event_type in event_types) {
  jc_file <- file.path(params$rmats_dir, paste0(event_type, ".MATS.JC.txt"))
  
  if (file.exists(jc_file)) {
    df <- read.delim(jc_file, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
      df$EventType <- event_type
      rmats_results[[event_type]] <- df
    }
  }
}

# Combine all events
if (length(rmats_results) > 0) {
  all_rmats <- bind_rows(rmats_results)
  
  # Add significance flag
  all_rmats <- all_rmats %>%
    mutate(
      Significant = FDR <= params$fdr_cutoff & abs(IncLevelDifference) >= params$dpsi_cutoff
    )
} else {
  all_rmats <- data.frame()
}
```

```{r rmats_summary, eval=!is.null(params$rmats_dir) && nrow(all_rmats) > 0}
# Summary table
summary_table <- all_rmats %>%
  group_by(EventType) %>%
  summarise(
    Total = n(),
    Significant = sum(Significant, na.rm = TRUE),
    FDR_leq_0.05 = sum(FDR <= 0.05, na.rm = TRUE),
    Mean_deltaPSI = mean(abs(IncLevelDifference), na.rm = TRUE)
  )

kable(summary_table, caption = "rMATS Event Summary by Event Type", digits = 3)
```

```{r rmats_volcano, eval=!is.null(params$rmats_dir) && nrow(all_rmats) > 0, fig.height=8}
# Volcano plot
all_rmats %>%
  filter(!is.na(FDR) & FDR > 0) %>%
  ggplot(aes(x = IncLevelDifference, y = -log10(FDR), color = Significant)) +
  geom_point(alpha = 0.5, size = 1) +
  facet_wrap(~EventType, scales = "free_y") +
  scale_color_manual(values = c("gray", "red")) +
  geom_hline(yintercept = -log10(params$fdr_cutoff), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = c(-params$dpsi_cutoff, params$dpsi_cutoff), linetype = "dashed", color = "blue") +
  theme_bw() +
  labs(
    title = "rMATS Volcano Plots",
    x = "ΔΨ (IncLevelDifference)",
    y = "-log10(FDR)"
  ) +
  theme(legend.position = "bottom")
```

```{r rmats_top_table, eval=!is.null(params$rmats_dir) && nrow(all_rmats) > 0}
# Top significant events
top_events <- all_rmats %>%
  filter(Significant) %>%
  arrange(FDR) %>%
  select(EventType, GeneID, geneSymbol, chr, strand, FDR, IncLevelDifference) %>%
  head(50)

if (nrow(top_events) > 0) {
  datatable(
    top_events,
    caption = "Top 50 Significant rMATS Events",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    formatRound(columns = c("FDR", "IncLevelDifference"), digits = 4)
} else {
  cat("No significant events detected.\n")
}
```

---

# MAJIQ Results

```{r majiq_load, eval=!is.null(params$majiq_dir)}
# Load MAJIQ TSV
tsv_file <- list.files(params$majiq_dir, pattern = "\\.tsv$", full.names = TRUE)[1]

if (!is.null(tsv_file) && file.exists(tsv_file)) {
  majiq_results <- read.delim(tsv_file, stringsAsFactors = FALSE)
  
  # Add significance flag (probability_changing >= 0.95)
  if ("probability_changing" %in% colnames(majiq_results)) {
    majiq_results <- majiq_results %>%
      mutate(Significant = probability_changing >= 0.95)
  } else {
    majiq_results$Significant <- FALSE
  }
} else {
  majiq_results <- data.frame()
}
```

```{r majiq_summary, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0}
cat("Total LSVs detected:", nrow(majiq_results), "\n\n")

if ("Significant" %in% colnames(majiq_results)) {
  cat("Significant LSVs (P(changing) >= 0.95):", sum(majiq_results$Significant, na.rm = TRUE), "\n\n")
}
```

```{r majiq_distribution, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0 && "mean_dpsi_per_lsv_junction" %in% colnames(majiq_results)}
# Distribution of deltaPSI
majiq_results %>%
  filter(!is.na(mean_dpsi_per_lsv_junction)) %>%
  ggplot(aes(x = abs(mean_dpsi_per_lsv_junction))) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  theme_bw() +
  labs(
    title = "Distribution of |ΔΨ| in MAJIQ LSVs",
    x = "|mean ΔΨ per LSV junction|",
    y = "Count"
  )
```

```{r majiq_top_table, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0}
# Top significant LSVs
cols_to_show <- intersect(
  c("gene_name", "lsv_id", "mean_dpsi_per_lsv_junction", "probability_changing", "probability_non_changing"),
  colnames(majiq_results)
)

if (length(cols_to_show) > 0) {
  top_lsvs <- majiq_results %>%
    filter(Significant) %>%
    arrange(desc(probability_changing)) %>%
    select(all_of(cols_to_show)) %>%
    head(50)
  
  if (nrow(top_lsvs) > 0) {
    datatable(
      top_lsvs,
      caption = "Top 50 Significant MAJIQ LSVs",
      options = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      formatRound(columns = names(top_lsvs)[sapply(top_lsvs, is.numeric)], digits = 4)
  } else {
    cat("No significant LSVs detected.\n")
  }
}
```

---

# IsoformSwitchAnalyzeR Results

```{r isar_load, eval=!is.null(params$isar_dir)}
# Load ISAR results
switches_file <- file.path(params$isar_dir, "top_isoform_switches.csv")
consequences_file <- file.path(params$isar_dir, "consequence_summary.csv")

if (file.exists(switches_file)) {
  isar_switches <- read.csv(switches_file, stringsAsFactors = FALSE)
} else {
  isar_switches <- data.frame()
}

if (file.exists(consequences_file)) {
  isar_consequences <- read.csv(consequences_file, stringsAsFactors = FALSE)
} else {
  isar_consequences <- data.frame()
}
```

```{r isar_summary, eval=!is.null(params$isar_dir) && nrow(isar_switches) > 0}
cat("Total isoform switches detected:", nrow(isar_switches), "\n\n")

if ("gene_name" %in% colnames(isar_switches)) {
  cat("Genes with switches:", length(unique(isar_switches$gene_name)), "\n\n")
}
```

```{r isar_consequences_plot, eval=!is.null(params$isar_dir) && nrow(isar_consequences) > 0, fig.height=6}
# Consequence summary plot
if ("feature" %in% colnames(isar_consequences) && "count" %in% colnames(isar_consequences)) {
  isar_consequences %>%
    ggplot(aes(x = reorder(feature, count), y = count)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    theme_bw() +
    labs(
      title = "Functional Consequences of Isoform Switches",
      x = "Consequence Type",
      y = "Count"
    )
}
```

```{r isar_top_table, eval=!is.null(params$isar_dir) && nrow(isar_switches) > 0}
# Top switches table
cols_to_show <- intersect(
  c("gene_name", "isoform_id", "condition_1", "condition_2", "dIF", "isoform_switch_q_value"),
  colnames(isar_switches)
)

if (length(cols_to_show) > 0) {
  top_switches <- isar_switches %>%
    head(50) %>%
    select(all_of(cols_to_show))
  
  datatable(
    top_switches,
    caption = "Top 50 Isoform Switches",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    formatRound(columns = names(top_switches)[sapply(top_switches, is.numeric)], digits = 4)
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
