---
title: "Alternative Splicing Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
params:
  comparison_id: "comparison"
  rmats_dir: NULL
  majiq_dir: NULL
  isar_dir: NULL
  sashimi_dir: NULL
  pegasas_dir: NULL
  leafcutter_dir: NULL
  fdr_cutoff: 0.05
  dpsi_cutoff: 0.1
  majiq_prob_threshold: 0.95
  majiq_dpsi_cutoff: 0.2
  sashimi_png_dpi: 150
  organism: "human"
  de_results: NULL
  run_de_as: false
  publish_deliverables: false
  publish_level: "core_raw"
  results_contract_version: "2.1.0"
  report_min_reads: 20
  group1_name: "Group1"
  group2_name: "Group2"
  tool_reliability:
    value: !r list()
  group1_sample_ids:
    value: !r character()
  group2_sample_ids:
    value: !r character()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(knitr)
  library(DT)
  library(ggrepel)
  library(UpSetR)
})

# Initialize result containers; populated by tool-specific chunks below.
# Cross-tool and enrichment sections rely on these always existing.
all_rmats         <- data.frame()
majiq_results     <- data.frame()
isar_switches     <- data.frame()
isar_consequences <- data.frame()
leafcutter_clusters <- data.frame()

# Replication sizes per group (empty arrays when a tool is disabled).
n1_samples <- length(as.character(unlist(params$group1_sample_ids)))
n2_samples <- length(as.character(unlist(params$group2_sample_ids)))
low_rep    <- (n1_samples > 0 && n1_samples < 4) ||
              (n2_samples > 0 && n2_samples < 4)

# Condition labels from comparisons.csv (group1/group2 columns). Fall back to
# Group1/Group2 when the pipeline does not pass names (e.g. manual renders).
cond_label <- function(name, fallback) {
  name <- as.character(name)
  if (length(name) == 0 || is.na(name) || !nzchar(name)) fallback else name
}
g1_label <- cond_label(params$group1_name, "Group1")
g2_label <- cond_label(params$group2_name, "Group2")

wrap_axis_labels <- function(x, width = 55) {
  vapply(x, function(s) paste(strwrap(s, width = width), collapse = "\n"), character(1))
}

# Reliability record per tool, shared with the deliverables manifests
# (contrast_manifest.yaml / run_manifest.yaml known_issues). Rendering is
# defensive: jsonlite can deliver the record as a data.frame (array of
# objects) or a list of named lists, and params may be absent entirely.
tool_issues_frame <- function() {
  rec <- params$tool_reliability
  if (is.null(rec)) return(NULL)
  if (is.data.frame(rec)) return(rec)
  if (is.list(rec) && length(rec) > 0 &&
      all(sapply(rec, function(item) is.list(item) || is.data.frame(item)))) {
    # list of named lists -> data.frame
    return(do.call(rbind, lapply(rec, function(item) {
      as.data.frame(lapply(item, function(x) if (is.null(x)) '' else x), stringsAsFactors = FALSE)
    })))
  }
  NULL
}

tool_issue <- function(tool) {
  frame <- tool_issues_frame()
  if (is.null(frame)) return(NULL)
  hit <- which(frame$tool == tool)
  if (length(hit) == 0) return(NULL)
  frame[hit[1], ]
}

tool_issue_status <- function(tool) {
  issue <- tool_issue(tool)
  if (is.null(issue)) return('ok')
  value <- as.character(issue$status)
  if (length(value) == 0 || is.na(value) || value == '') return('unknown')
  value
}

tool_issue_note <- function(tool) {
  issue <- tool_issue(tool)
  if (is.null(issue)) return('')
  value <- as.character(issue$note)
  if (length(value) == 0 || is.na(value)) return('')
  value
}

style_ora_dotplot <- function(p, title) {
  p +
    labs(title = title) +
    scale_y_discrete(labels = function(lbl) wrap_axis_labels(lbl, width = 55)) +
    theme_bw(base_size = 11) +
    theme(
      axis.text.y = element_text(size = 8, lineheight = 0.9),
      plot.margin = margin(5.5, 5.5, 5.5, 25)
    )
}
```

# Summary

This report presents differential alternative splicing analysis results for comparison: **`r params$comparison_id`**.

**Analysis tools:**

- rMATS-turbo: Differential alternative splicing (SE, A5SS, A3SS, MXE, RI events)
- MAJIQ V3: Local Splicing Variation (LSV) detection
- IsoformSwitchAnalyzeR: Isoform switch identification with functional consequences

**Filter criteria:**

- rMATS / ISAR FDR cutoff: `r params$fdr_cutoff`
- rMATS / ISAR |ΔΨ| / |dIF| cutoff: `r params$dpsi_cutoff`
- MAJIQ P(|ΔΨ| > threshold) cutoff: `r params$majiq_prob_threshold`
- MAJIQ |ΔΨ| cutoff: `r params$majiq_dpsi_cutoff`

```{r summary_warnings, results='asis'}
if (low_rep) {
  cat(sprintf(
    "**Low replication warning:** this comparison has n=%d (%s) vs n=%d (%s). ",
    n1_samples, g1_label, n2_samples, g2_label
  ))
  cat(
    "Statistical power for detecting moderate ΔΨ is limited; ",
    "interpret borderline calls with caution.\n\n"
  )
}
```

```{r known_issues_summary, results='asis'}
# Surface the manifest tool_reliability record early — users of the tables
# below (or the deliverables layer) see the caveats before any tool section.
issues <- tool_issues_frame()
if (!is.null(issues) && nrow(issues) > 0) {
  bad <- issues[is.na(issues$status) | issues$status != 'ok', ]
  if (nrow(bad) > 0) {
    cat("<hr>\n\n**Known tool reliability issues (from the run manifest):**\n\n")
    for (i in seq_len(nrow(bad))) {
      note <- as.character(bad$note[i])
      if (length(note) == 0 || is.na(note) || note == '') note <- ''
      cat(sprintf(
        "<p style='color:#b30000'><strong>%s</strong> [%s]: %s</p>\n",
        as.character(bad$tool[i]), as.character(bad$status[i]), note
      ))
    }
    cat("\n")
  }
}
```

---

# rMATS Results

```{r rmats_load, eval=!is.null(params$rmats_dir)}
event_types <- c("SE", "A5SS", "A3SS", "MXE", "RI")
rmats_results_list <- list()

for (event_type in event_types) {
  jc_file <- file.path(params$rmats_dir, paste0(event_type, ".MATS.JC.txt"))
  if (file.exists(jc_file)) {
    # De novo (unannotated) splice sites: rMATS 4.3 no longer encodes them as
    # negative coordinates, so match event IDs against fromGTF.novelSpliceSite.
    # The rMATS 4.3 SE header duplicates the ID column; df$ID is always the
    # first (event) ID because read.delim renames the duplicate to ID.1.
    novel_ids <- character(0)
    novel_file <- file.path(params$rmats_dir,
                            paste0("fromGTF.novelSpliceSite.", event_type, ".txt"))
    if (file.exists(novel_file)) {
      novel_ids <- read.delim(novel_file, stringsAsFactors = FALSE)$ID
    }

    df <- read.delim(jc_file, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
      df$EventType <- event_type
      df$NovelSite <- df$ID %in% novel_ids
      rmats_results_list[[event_type]] <- df
    }
  }
}

if (length(rmats_results_list) > 0) {
  # Sum of a comma-separated per-sample count vector (0 when not parseable).
  parse_count_sum <- function(x) {
    if (is.na(x) || x == "" || x == "NA") return(0)
    vals <- suppressWarnings(as.numeric(strsplit(as.character(x), ",")[[1]]))
    vals <- vals[!is.na(vals)]
    if (length(vals) == 0) return(0)
    sum(vals)
  }

  # 0 disables the ranking-table coverage filter (e.g. manual renders without
  # the param in their JSON).
  min_reads <- params$report_min_reads
  if (is.null(min_reads)) min_reads <- 0

  all_rmats <- bind_rows(rmats_results_list) %>%
    mutate(
      FDR_plot       = pmax(FDR, .Machine$double.xmin),
      # Cap -log10(FDR): underflowed --cstat p-values would otherwise push the
      # priority score towards a single saturated value; past 1e-50 the ranking
      # is effectively driven by |dPsi|.
      neglog10FDR    = pmin(-log10(FDR_plot), 50),
      Significant    = FDR <= params$fdr_cutoff & abs(IncLevelDifference) >= params$dpsi_cutoff,
      # Higher = more significant and larger PSI change.
      priority_score = round(neglog10FDR * abs(IncLevelDifference), 3),
      # Splice-site class: reflects splice-site novelty only.
      #   annotated = every splice site present in the GTF annotation (events
      #               with known sites but a novel junction stay here);
      #   de_novo   = at least one site absent from it (fromGTF.novelSpliceSite).
      site_class     = ifelse(NovelSite, "de_novo", "annotated"),
      # Junction-read support per group (IJC + SJC). Drives the report_min_reads
      # filter of the ranking tables only; significance calling is untouched.
      reads_group1   = mapply(parse_count_sum, IJC_SAMPLE_1) +
                       mapply(parse_count_sum, SJC_SAMPLE_1),
      reads_group2   = mapply(parse_count_sum, IJC_SAMPLE_2) +
                       mapply(parse_count_sum, SJC_SAMPLE_2),
      reads_ok       = reads_group1 >= min_reads & reads_group2 >= min_reads
    )
}
```

```{r rmats_panel_prep, eval=!is.null(params$rmats_dir) && exists("all_rmats") && nrow(all_rmats) > 0}
# ---- shared helpers for the three rMATS panels (Combined / Annotated / De novo) ----

parse_psi_vec <- function(x) {
  if (is.na(x) || x == "NA" || x == "") return(NA_real_)
  suppressWarnings(as.numeric(strsplit(as.character(x), ",")[[1]]))
}

resolve_sample_labels <- function(expected_n, ids, prefix) {
  ids <- as.character(unlist(ids))
  ids <- ids[!is.na(ids) & ids != ""]
  if (length(ids) == expected_n) {
    return(ids)
  }
  # Sample ids are now guaranteed to match the rMATS column order by
  # construction; a count mismatch means a real mapping problem, not a case
  # for silent placeholder labels.
  warning(sprintf(
    "Sample-id count (%d) does not match rMATS column count (%d) for group %s — ",
    length(ids), expected_n, prefix
  ), "using placeholder labels. Check sample/condition mapping in the samplesheet.")
  paste0(prefix, "_rep", seq_len(expected_n))
}

parse_sample_counts <- function(count_str, group_label) {
  if (is.na(count_str) || count_str == "") return(NULL)
  vals <- suppressWarnings(as.integer(strsplit(as.character(count_str), ",")[[1]]))
  if (length(vals) == 0) return(NULL)
  data.frame(
    sample_idx = seq_along(vals),
    count      = vals,
    group      = group_label,
    stringsAsFactors = FALSE
  )
}

# Non-significant points per volcano: all significant events are always drawn,
# the background is subsampled to keep the HTML at a sane size.
RMATS_NONSIG_PLOT_MAX <- 20000

panel_summary_table <- function(df) {
  df %>%
    group_by(EventType) %>%
    summarise(
      Total       = n(),
      Significant = sum(Significant, na.rm = TRUE),
      FDR_leq_0.05 = sum(FDR <= 0.05, na.rm = TRUE),
      # Mean |dPsi| over ALL events of the type, not only the significant ones.
      Mean_abs_deltaPSI_all_events = round(mean(abs(IncLevelDifference), na.rm = TRUE), 3),
      # Significant events clearing the ranking-table coverage minimum.
      Significant_min_reads = sum(Significant & reads_ok, na.rm = TRUE)
    ) %>%
    kable(caption = paste(
      "rMATS Event Summary by Event Type.",
      "Mean_abs_deltaPSI_all_events is the mean |deltaPSI| across all events of",
      "the type (significant or not), so it sits well below the significance cutoff.",
      "Significant_min_reads counts significant events with >= report_min_reads",
      "junction reads (IJC+SJC per group); it is a ranking-table filter, never a",
      "significance one."
    ))
}

panel_volcano <- function(df, label) {
  # Top-5 labels per type, tie-broken by junction-read coverage: past the
  # -log10(FDR) cap of 50 the priority score saturates and |dPsi| alone would
  # decide the order.
  top_labels <- df %>%
    filter(Significant, reads_ok,
           !is.na(geneSymbol), geneSymbol != "",
           !is.na(IncLevelDifference), !is.na(neglog10FDR)) %>%
    mutate(cov_min = pmin(reads_group1, reads_group2)) %>%
    arrange(desc(priority_score), desc(cov_min)) %>%
    group_by(EventType) %>%
    slice_head(n = 5) %>%
    ungroup()

  plot_df <- df %>% filter(!is.na(IncLevelDifference), !is.na(neglog10FDR))
  sig     <- plot_df %>% filter(Significant)
  nonsig  <- plot_df %>% filter(!Significant)
  n_nonsig_orig <- nrow(nonsig)
  if (n_nonsig_orig > RMATS_NONSIG_PLOT_MAX) {
    set.seed(1)
    nonsig <- dplyr::sample_n(nonsig, RMATS_NONSIG_PLOT_MAX)
  }
  sub_note <- if (n_nonsig_orig > RMATS_NONSIG_PLOT_MAX) {
    sprintf(" (%d of %d non-significant events shown)", RMATS_NONSIG_PLOT_MAX, n_nonsig_orig)
  } else NULL

  bind_rows(sig, nonsig) %>%
    ggplot(aes(x = IncLevelDifference, y = neglog10FDR, color = Significant)) +
    geom_point(alpha = 0.4, size = 1) +
    geom_text_repel(
      data = top_labels,
      aes(y = neglog10FDR, label = geneSymbol),
      size = 2.5, max.overlaps = 10, color = "black"
    ) +
    facet_wrap(~EventType, scales = "free_y") +
    scale_color_manual(values = c("gray70", "firebrick")) +
    geom_hline(yintercept = -log10(params$fdr_cutoff),    linetype = "dashed", color = "navy", linewidth = 0.4) +
    geom_vline(xintercept = c(-params$dpsi_cutoff, params$dpsi_cutoff), linetype = "dashed", color = "navy", linewidth = 0.4) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
    coord_cartesian(clip = "off") +
    theme_bw() +
    labs(
      title    = paste0("rMATS Volcano Plots (", label, ")"),
      subtitle = sub_note,
      x        = "ΔΨ (IncLevelDifference)",
      y        = "-log10(FDR), capped at 50"
    ) +
    theme(
      legend.position = "bottom",
      plot.margin = margin(t = 5.5, r = 35, b = 5.5, l = 5.5)
    )
}

panel_top_table <- function(df, label, top_n, element_id) {
  top_events <- df %>%
    filter(Significant, reads_ok) %>%
    # Tie-break saturated priority scores (past the -log10(FDR) cap) by the
    # weakest group's junction-read coverage.
    arrange(desc(priority_score), desc(pmin(reads_group1, reads_group2))) %>%
    select(EventType, GeneID, geneSymbol, chr, strand, FDR, IncLevelDifference,
           priority_score, site_class, reads_group1, reads_group2) %>%
    head(top_n) %>%
    mutate(
      # Keep underflowed --cstat p-values readable instead of rounding them to 0.
      FDR_display = ifelse(FDR == 0, "< 1e-300", formatC(FDR, format = "e", digits = 2))
    ) %>%
    select(-FDR)
  if (nrow(top_events) == 0) {
    cat("No significant events detected in this panel.\n\n")
    return(invisible(NULL))
  }
  cat(sprintf(
    "Ranking filtered to events with >= %s junction reads (IJC+SJC) per group (`report_min_reads`; significance calling is unaffected).\n\n",
    min_reads
  ))
  print(
    datatable(
      top_events,
      elementId = element_id,
      caption   = sprintf("Top %d significant rMATS events (%s) ranked by priority score",
                          top_n, label),
      options   = list(pageLength = 10, scrollX = TRUE),
      rownames  = FALSE, escape = FALSE
    ) %>%
      formatRound(columns = c("IncLevelDifference", "priority_score"), digits = 4) %>%
      formatRound(columns = c("reads_group1", "reads_group2"), digits = 0)
  )
}

panel_pca <- function(df, label) {
  pca_event_type <- if ("SE" %in% df$EventType) "SE" else {
    df %>% count(EventType) %>% slice_max(n, n = 1) %>% pull(EventType)
  }
  # Coverage filter on the PCA event pool too: variance ranking on uncovered
  # de novo events would otherwise visualise noise (report_min_reads).
  pca_events_raw <- if (length(pca_event_type) > 0) {
    df %>%
      filter(reads_ok,
             EventType == pca_event_type[1],
             !is.na(IncLevel1), !is.na(IncLevel2),
             IncLevel1 != "NA", IncLevel2 != "NA")
  } else data.frame()

  if (nrow(pca_events_raw) < 10) {
    cat(sprintf("Insufficient events with valid PSI data for the %s PCA (need >= 10).\n\n", label))
    return(invisible(NULL))
  }

  psi_all <- lapply(seq_len(nrow(pca_events_raw)), function(i) {
    c(parse_psi_vec(pca_events_raw$IncLevel1[i]),
      parse_psi_vec(pca_events_raw$IncLevel2[i]))
  })
  n_samples <- length(psi_all[[1]])
  valid_events <- sapply(psi_all, function(v) {
    length(v) == n_samples && sum(!is.na(v)) >= 3
  })
  psi_all <- psi_all[valid_events]
  if (length(psi_all) < 10) {
    cat(sprintf("Insufficient events with valid PSI data for the %s PCA (need >= 10).\n\n", label))
    return(invisible(NULL))
  }

  event_variance <- sapply(psi_all, function(v) var(v, na.rm = TRUE))
  keep_idx <- order(event_variance, decreasing = TRUE)[seq_len(min(2000, length(psi_all)))]
  psi_list <- psi_all[keep_idx]

  psi_mat <- do.call(rbind, psi_list)
  psi_mat <- apply(psi_mat, 1, function(row) {
    med <- median(row, na.rm = TRUE)
    row[is.na(row)] <- med
    row
  }) %>% t()

  keep_row <- which(valid_events)[keep_idx[1]]
  n1 <- length(parse_psi_vec(pca_events_raw$IncLevel1[keep_row]))
  n2 <- length(parse_psi_vec(pca_events_raw$IncLevel2[keep_row]))

  sample_labels <- c(
    resolve_sample_labels(n1, params$group1_sample_ids, g1_label),
    resolve_sample_labels(n2, params$group2_sample_ids, g2_label)
  )
  group_labels  <- c(rep(g1_label, n1), rep(g2_label, n2))

  pca_res <- tryCatch(
    prcomp(t(psi_mat), center = TRUE, scale. = TRUE),
    error = function(e) NULL
  )
  if (is.null(pca_res)) return(invisible(NULL))

  var_exp <- round(summary(pca_res)$importance[2, 1:min(2, ncol(pca_res$x))] * 100, 1)
  pca_df  <- as.data.frame(pca_res$x[, 1:min(2, ncol(pca_res$x))])
  pca_df$Group  <- group_labels
  pca_df$Sample <- sample_labels

  print(
    ggplot(pca_df, aes(x = PC1, y = PC2, color = Group, label = Sample)) +
      geom_point(size = 3) +
      geom_text_repel(size = 3, max.overlaps = 20) +
      theme_bw() +
      scale_color_brewer(palette = "Set1") +
      labs(
        title = paste0("PCA of PSI profiles (", label, ": ", pca_event_type[1],
                       " events, n=", length(psi_list), ", per-event median imputation, scaled)"),
        x = paste0("PC1 (", var_exp[1], "% variance)"),
        y = paste0("PC2 (", var_exp[2], "% variance)")
      )
  )
}

panel_coverage <- function(df, label) {
  jq_pool <- df %>%
    filter(EventType == "SE", !is.na(IJC_SAMPLE_1), !is.na(IJC_SAMPLE_2))
  if (nrow(jq_pool) == 0) {
    cat(sprintf("No SE events with junction count data available for the %s panel.\n\n", label))
    return(invisible(NULL))
  }
  # Fixed-seed random sample (not head()): the first rows in file order are an
  # arbitrary, non-representative slice of the event list.
  set.seed(1)
  jq_events <- if (nrow(jq_pool) > 500) dplyr::sample_n(jq_pool, 500) else jq_pool
  jq_raw <- bind_rows(lapply(seq_len(nrow(jq_events)), function(i) {
    bind_rows(
      parse_sample_counts(jq_events$IJC_SAMPLE_1[i], g1_label),
      parse_sample_counts(jq_events$IJC_SAMPLE_2[i], g2_label)
    )
  }))
  if (nrow(jq_raw) == 0) return(invisible(NULL))

  jq_summary <- jq_raw %>%
    group_by(group, sample_idx) %>%
    summarise(mean_IJC = mean(count, na.rm = TRUE), .groups = "drop")

  g1_ids <- resolve_sample_labels(
    sum(jq_summary$group == g1_label), params$group1_sample_ids, g1_label
  )
  g2_ids <- resolve_sample_labels(
    sum(jq_summary$group == g2_label), params$group2_sample_ids, g2_label
  )
  jq_summary <- jq_summary %>%
    group_by(group) %>%
    arrange(sample_idx, .by_group = TRUE) %>%
    mutate(Sample = if_else(group == g1_label, g1_ids[sample_idx], g2_ids[sample_idx])) %>%
    ungroup()

  print(
    ggplot(jq_summary, aes(x = reorder(Sample, mean_IJC), y = mean_IJC, fill = group)) +
      geom_col(alpha = 0.8) +
      coord_flip() +
      theme_bw() +
      scale_fill_brewer(palette = "Set1") +
      labs(
        title    = paste0("Mean inclusion junction count per sample (", label, ")"),
        subtitle = sprintf(
          "Random sample of %d SE events with junction counts (seed 1); low values may indicate insufficient coverage",
          nrow(jq_events)
        ),
        x = NULL,
        y = "Mean IJC"
      )
  )
}

sashimi_widget_html <- function(widget_id) {
  paste0(
    '<div id="', widget_id, '" style="border:1px solid #d9d9d9; border-radius:8px; padding:12px; margin:12px 0 20px 0;">',
    '<div style="font-weight:600; margin-bottom:8px;">Sashimi browser</div>',
    '<div style="display:flex; flex-wrap:wrap; gap:10px; align-items:flex-end; margin-bottom:10px;">',
    '<label style="font-size:0.9em;">Event type<br><select id="', widget_id, '_event"></select></label>',
    '<label style="font-size:0.9em;">Gene<br><select id="', widget_id, '_gene"></select></label>',
    '</div>',
    '<div id="', widget_id, '_holder" style="width:100%; height:760px; border:1px solid #e5e5e5; border-radius:6px; overflow:auto; background:#fafafa;">',
    '<p style="padding:10px;">Select an event type and gene above to preview the sashimi plot.</p>',
    '</div>',
    '</div>'
  )
}

panel_sashimi <- function(class_filter, widget_suffix) {
  if (!exists("sashimi_table") || nrow(sashimi_table) == 0) {
    cat("Sashimi plots not available for this run.\n\n")
    return(invisible(NULL))
  }
  widget_id <- paste0("sashimi_browser_", widget_suffix)
  cat(sashimi_widget_html(widget_id))
  cat('\n<script>window.__sashimiBrowser("', widget_id, '", ',
      if (is.null(class_filter)) "null" else paste0('"', class_filter, '"'),
      ');</script>\n\n', sep = "")
}

render_rmats_panel <- function(panel_df, label, class_note, widget_suffix,
                               class_filter, top_n) {
  n_total <- nrow(panel_df)
  n_sig   <- sum(panel_df$Significant, na.rm = TRUE)
  cat(sprintf(
    "**%s panel**: %d events; %d significant (FDR <= %s & |ΔΨ| >= %s).\n\n",
    label, n_total, n_sig, params$fdr_cutoff, params$dpsi_cutoff
  ))
  if (n_total == 0) {
    cat("No events in this panel.\n")
    return(invisible(NULL))
  }
  cat(sprintf("*%s*\n\n", class_note))
  cat("#### Event summary\n\n")
  print(panel_summary_table(panel_df))
  cat("\n\n#### Volcano\n\n")
  print(panel_volcano(panel_df, label))
  cat("\n\n#### Top significant events\n\n")
  panel_top_table(panel_df, label, top_n, paste0("rmats_top_", widget_suffix))
  cat("\n\n#### PSI profile (PCA)\n\n")
  panel_pca(panel_df, label)
  cat("\n\n#### Splice junction coverage\n\n")
  panel_coverage(panel_df, label)
  cat("\n\n#### Sashimi plots\n\n")
  panel_sashimi(class_filter, widget_suffix)
  invisible(NULL)
}

rmats_annotated <- all_rmats %>% filter(site_class == "annotated")
rmats_denovo    <- all_rmats %>% filter(site_class == "de_novo")
```

```{r sashimi_shared_data, eval=!is.null(params$sashimi_dir), results='asis'}
# Embed every sashimi PDF once as a data URI; the three panel browsers reuse
# this shared payload client-side (no per-panel duplication of the base64 blob).
sashimi_pdfs <- list.files(params$sashimi_dir, pattern = "\\.pdf$",
                           recursive = TRUE, full.names = TRUE)

sashimi_table <- data.frame()

if (length(sashimi_pdfs) > 0) {
  event_from_path <- function(paths) {
    rel <- basename(dirname(dirname(paths)))
    ifelse(rel %in% c("SE", "A5SS", "A3SS", "MXE", "RI"), rel, NA_character_)
  }

  extract_gene_symbol <- function(x) {
    has_prefix <- grepl("^[0-9]+_[^_]+_", x)
    out <- ifelse(has_prefix, sub("^[0-9]+_([^_]+)_.*$", "\\1", x), NA_character_)
    out
  }

  # Convert each PDF to a PNG (~150 dpi) so the HTML stays light and
  # self-contained. Falls back to embedding the PDF when pdftoppm is missing
  # (older report container).
  has_pdftoppm <- nzchar(Sys.which("pdftoppm"))

  embed_sashimi <- function(pdf_path) {
    pdf_size <- file.info(pdf_path)$size
    if (is.na(pdf_size) || pdf_size <= 0) return(NA_character_)

    if (has_pdftoppm) {
      tmpdir <- tempfile("sashimi_png_")
      dir.create(tmpdir, showWarnings = FALSE)
      out_stem <- file.path(tmpdir, "plot")
      status <- system2("pdftoppm", c("-png", "-r", as.character(params$sashimi_png_dpi),
                                      shQuote(pdf_path), shQuote(out_stem)),
                        stdout = FALSE, stderr = FALSE)
      png_files <- list.files(tmpdir, pattern = "\\.png$", full.names = TRUE)
      if (status == 0 && length(png_files) > 0) {
        png_size <- file.info(png_files[1])$size
        raw_png  <- readBin(png_files[1], what = "raw", n = png_size)
        return(paste0("data:image/png;base64,", jsonlite::base64_enc(raw_png)))
      }
    }

    # Fallback: embed the vector PDF directly
    raw_pdf <- readBin(pdf_path, what = "raw", n = pdf_size)
    paste0("data:application/pdf;base64,", jsonlite::base64_enc(raw_pdf))
  }

  sashimi_data_uris <- vapply(sashimi_pdfs, embed_sashimi, character(1))

  sashimi_table <- data.frame(
    file = basename(sashimi_pdfs),
    event_type = event_from_path(sashimi_pdfs),
    gene = extract_gene_symbol(basename(sashimi_pdfs)),
    # rmats2sashimiplot cannot draw unannotated splice sites, so
    # filter_rmats_for_sashimi.py excludes de novo events upstream and every
    # generated plot is annotated. The column keeps the browser contract ready
    # for a per-class split if plotting ever supports de novo sites.
    site_class = "annotated",
    data_uri = sashimi_data_uris,
    mime = ifelse(grepl("^data:image/png", sashimi_data_uris), "image/png", "application/pdf"),
    stringsAsFactors = FALSE
  ) %>%
    dplyr::arrange(event_type, gene, file) %>%
    dplyr::filter(!is.na(data_uri) & nzchar(data_uri))
}

if (nrow(sashimi_table) > 0) {
  browser_json <- jsonlite::toJSON(
    sashimi_table[, c("file", "event_type", "gene", "site_class", "data_uri", "mime")],
    dataframe = "rows", auto_unbox = TRUE
  )
  cat('<script>window.__SASHIMI_PLOTS__ = ', browser_json, ';</script>\n')
  cat(paste0(
    '<script>(function(){',
    'window.__sashimiBrowser = function (widgetId, classFilter) {',
    'const root = document.getElementById(widgetId);',
    'if (!root) { return; }',
    'const eventSel = document.getElementById(widgetId + "_event");',
    'const geneSel = document.getElementById(widgetId + "_gene");',
    'const holder = document.getElementById(widgetId + "_holder");',
    'const all = Array.isArray(window.__SASHIMI_PLOTS__) ? window.__SASHIMI_PLOTS__ : [];',
    'const data = classFilter ? all.filter(d => d.site_class === classFilter) : all;',
    'const uniq = (arr) => Array.from(new Set(arr));',
    'const sortText = (arr) => arr.sort((a,b) => String(a).localeCompare(String(b)));',
    'const setOptions = (sel, values) => {',
    'sel.innerHTML = "";',
    'values.forEach(v => { const o=document.createElement("option"); o.value=v; o.textContent=v; sel.appendChild(o); });',
    '};',
    'const renderPreview = (row) => {',
    'holder.innerHTML = "";',
    'if(!row){ holder.innerHTML = "<p style=\\"padding:10px;\\">No plot available.</p>"; return; }',
    'if(row.mime === "image/png"){',
    'const img = document.createElement("img");',
    'img.src = row.data_uri;',
    'img.style.maxWidth = "100%";',
    'img.style.display = "block";',
    'img.style.margin = "0 auto";',
    'holder.appendChild(img);',
    '} else {',
    'const obj = document.createElement("object");',
    'obj.type = "application/pdf";',
    'obj.data = row.data_uri;',
    'obj.style.width = "100%";',
    'obj.style.height = "760px";',
    'holder.appendChild(obj);',
    '}',
    '};',
    'if (data.length === 0) {',
    'holder.innerHTML = "<p style=\\"padding:10px;\\">No sashimi plots available for this class: rmats2sashimiplot cannot draw unannotated (de novo) splice sites, so sashimi output covers annotated events only.</p>";',
    'return;',
    '}',
    'const eventValues = sortText(uniq(data.map(d => d.event_type)));',
    'setOptions(eventSel, eventValues);',
    'const currentRows = () => data.filter(d => d.event_type === eventSel.value);',
    'const refreshGenes = () => {',
    'const genes = sortText(uniq(currentRows().map(d => d.gene)));',
    'setOptions(geneSel, genes);',
    '};',
    'const refreshPreview = () => {',
    'const row = currentRows().find(d => d.gene === geneSel.value);',
    'renderPreview(row);',
    '};',
    'eventSel.addEventListener("change", () => { refreshGenes(); refreshPreview(); });',
    'geneSel.addEventListener("change", refreshPreview);',
    'refreshGenes(); refreshPreview();',
    '};',
    '})();</script>'
  ))
}
```

## rMATS event panels {.tabset .tabset-pills}

### Combined {.active}

```{r rmats_panel_combined, eval=!is.null(params$rmats_dir) && exists("all_rmats") && nrow(all_rmats) > 0, results='asis', fig.width=10, fig.height=7}
render_rmats_panel(
  all_rmats, "Combined",
  "All events: annotated and de novo splice-site classes together.",
  "combined", NULL, 100
)
```

### Annotated splice sites

```{r rmats_panel_annotated, eval=!is.null(params$rmats_dir) && exists("all_rmats") && nrow(all_rmats) > 0, results='asis', fig.width=10, fig.height=7}
render_rmats_panel(
  rmats_annotated, "Annotated splice sites",
  "Every splice site present in the GTF annotation. Note: the class reflects splice-site novelty only — events with known splice sites but a novel junction (fromGTF.novelJunction) also fall here.",
  "annotated", "annotated", 100
)
```

### De novo splice sites

```{r rmats_panel_denovo, eval=!is.null(params$rmats_dir) && exists("all_rmats") && nrow(all_rmats) > 0, results='asis', fig.width=10, fig.height=7}
render_rmats_panel(
  rmats_denovo, "De novo splice sites",
  "At least one splice site absent from the GTF annotation (fromGTF.novelSpliceSite). Coverage-dependent by nature: check the coverage plot and the Significant_min_reads column before interpreting the ranking.",
  "de_novo", "de_novo", 200
)
```


Vector PDFs for publication are published inside the deliverables folder, under
`contrasts/<comparison_id>/plots/sashimi/<EVENT_TYPE>/`, and indexed by
`contrasts/<comparison_id>/plots/sashimi/sashimi_index.tsv`.


---

# MAJIQ Results

```{r majiq_load, eval=!is.null(params$majiq_dir)}
tsv_files <- list.files(params$majiq_dir, pattern = "\\.tsv$", full.names = TRUE)
tsv_file  <- if (length(tsv_files) > 0) tsv_files[1] else NA_character_

if (!is.na(tsv_file) && file.exists(tsv_file)) {
  majiq_results <- tryCatch(
    # comment.char skips the JSON metadata block (#-prefixed lines) at the top of MAJIQ v3 TSVs
    read.delim(tsv_file, stringsAsFactors = FALSE, comment.char = "#"),
    error = function(e) data.frame()
  )

  if (nrow(majiq_results) > 0 &&
      "probability_changing" %in% colnames(majiq_results) &&
      "dpsi_mean" %in% colnames(majiq_results)) {
    majiq_results <- majiq_results %>%
      mutate(Significant = probability_changing >= params$majiq_prob_threshold &
               abs(dpsi_mean) >= params$majiq_dpsi_cutoff)
  } else {
    majiq_results$Significant <- FALSE
  }

  if (nrow(majiq_results) > 0) {
    if ("gene_name" %in% colnames(majiq_results)) {
      gene_name_clean <- trimws(as.character(majiq_results$gene_name))
      gene_id_clean <- if ("gene_id" %in% colnames(majiq_results)) {
        trimws(as.character(majiq_results$gene_id))
      } else {
        rep(NA_character_, nrow(majiq_results))
      }

      # Keep only symbol-like labels as informative genes; drop blanks and ID echoes.
      majiq_results$InformativeGene <- !is.na(gene_name_clean) & gene_name_clean != "" &
        (is.na(gene_id_clean) | gene_id_clean == "" | gene_name_clean != gene_id_clean)
    } else {
      majiq_results$InformativeGene <- TRUE
    }
  }
}
```

```{r majiq_summary, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0}
majiq_summary_rows <- data.frame(
  Metric = c(
    "Total junctions detected",
    sprintf("Significant LSVs (P(|ΔPSI|) >= %.2f & |ΔΨ| >= %.2f)",
            params$majiq_prob_threshold, params$majiq_dpsi_cutoff)
  ),
  Value = c(nrow(majiq_results), NA_integer_),
  stringsAsFactors = FALSE
)

if ("Significant" %in% colnames(majiq_results)) {
  majiq_summary_rows$Value[2] <- sum(majiq_results$Significant, na.rm = TRUE)

  if ("InformativeGene" %in% colnames(majiq_results)) {
    majiq_summary_rows <- rbind(
      majiq_summary_rows,
      data.frame(
        Metric = c("Significant LSVs with informative gene symbol",
                   "Significant LSVs without informative gene symbol"),
        Value = c(
          sum(majiq_results$Significant & majiq_results$InformativeGene, na.rm = TRUE),
          sum(majiq_results$Significant & !majiq_results$InformativeGene, na.rm = TRUE)
        ),
        stringsAsFactors = FALSE
      )
    )
  }
}

knitr::kable(majiq_summary_rows, caption = "MAJIQ summary")
```

```{r majiq_distribution, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0 && "dpsi_mean" %in% colnames(majiq_results)}
majiq_results %>%
  filter(!is.na(dpsi_mean)) %>%
  ggplot(aes(x = abs(dpsi_mean))) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  theme_bw() +
  labs(
    title = "Distribution of |ΔΨ| in MAJIQ junctions",
    x     = "|dpsi_mean|",
    y     = "Count"
  )
```

```{r majiq_top_table, eval=!is.null(params$majiq_dir) && nrow(majiq_results) > 0}
cols_to_show <- intersect(
  c("gene_name", "gene_id", "event_type", "seqid", "start", "end",
    "dpsi_mean", "dpsi_std", "probability_changing", "probability_nonchanging",
    "grp1_raw_psi_mean", "grp2_raw_psi_mean"),
  colnames(majiq_results)
)

if (length(cols_to_show) > 0) {
  top_lsvs <- majiq_results %>%
    filter(Significant)

  if ("InformativeGene" %in% colnames(top_lsvs)) {
    top_lsvs <- top_lsvs %>% filter(InformativeGene)
  }

  top_lsvs <- top_lsvs %>%
    arrange(desc(probability_changing), desc(abs(dpsi_mean))) %>%
    select(all_of(cols_to_show)) %>%
    head(100)

  if (nrow(top_lsvs) > 0) {
    datatable(
      top_lsvs,
      caption  = paste0("Top 100 significant MAJIQ junctions with informative gene symbols (P >= ",
                        params$majiq_prob_threshold, ", |ΔΨ| >= ",
                        params$majiq_dpsi_cutoff, ")"),
      options  = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      formatRound(columns = names(top_lsvs)[sapply(top_lsvs, is.numeric)], digits = 4)
  } else {
    cat("No significant junctions detected.\n")
  }
}
```

---

# IsoformSwitchAnalyzeR Results

```{r isar_load, eval=!is.null(params$isar_dir)}
read_csv_safe <- function(path) {
  if (!file.exists(path) || file.size(path) < 3L) return(data.frame())
  tryCatch(read.csv(path, stringsAsFactors = FALSE), error = function(e) data.frame())
}

isar_switches     <- read_csv_safe(file.path(params$isar_dir, "top_isoform_switches.csv"))
isar_consequences <- read_csv_safe(file.path(params$isar_dir, "consequence_summary.csv"))
```

```{r isar_summary, eval=!is.null(params$isar_dir) && nrow(isar_switches) > 0, results='asis'}
if ("isoform_switch_q_value" %in% colnames(isar_switches) &&
    "dIF" %in% colnames(isar_switches)) {
  isar_switches$significant <- (
    isar_switches$isoform_switch_q_value < params$fdr_cutoff &
    abs(isar_switches$dIF) >= params$dpsi_cutoff
  )
} else {
  isar_switches$significant <- NA
}

n_total <- nrow(isar_switches)
n_sig   <- sum(isar_switches$significant, na.rm = TRUE)

cat("Total isoforms tested:", n_total, "\n\n")
cat(sprintf("Significant (FDR < %.2f & |dIF| >= %.2f): %d\n\n",
            params$fdr_cutoff, params$dpsi_cutoff, n_sig))

if ("gene_name" %in% colnames(isar_switches)) {
  cat("Genes tested:", length(unique(isar_switches$gene_name)), "\n")
  if (n_sig > 0) {
    sig_genes_isar_n <- length(unique(isar_switches$gene_name[isar_switches$significant]))
    cat("Genes with significant switches:", sig_genes_isar_n, "\n")
  }
}

# Auto-warning for a saturated test: q-values pinned in a narrow band while
# large dIF exist often signals a satuRn convergence failure rather than a
# genuinely empty contrast (see plans/report_bugs_fix_plan.md, P1-1).
q_vals <- isar_switches$isoform_switch_q_value[!is.na(isar_switches$isoform_switch_q_value)]
d_vals <- isar_switches$dIF[!is.na(isar_switches$dIF)]
if (length(q_vals) > 10 && IQR(q_vals) < 0.05 && max(abs(d_vals)) >= 0.1) {
  cat(paste0(
    "<p style='color:#b30000'>WARNING: satuRn q-values are pinned in a narrow range ",
    "(IQR &lt; 0.05) while isoforms with |dIF| >= 0.1 exist. This usually indicates a ",
    "convergence/ordering failure of the test (see P1-1), not a genuinely empty contrast.</p>\n\n"
  ))
}
```

```{r isar_consequences_plot, eval=!is.null(params$isar_dir) && nrow(isar_consequences) > 0, fig.height=6}
if ("feature" %in% colnames(isar_consequences) && "count" %in% colnames(isar_consequences)) {
  isar_consequences %>%
    ggplot(aes(x = reorder(feature, count), y = count)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    theme_bw() +
    labs(
      title = "Functional Consequences of Isoform Switches",
      x     = "Consequence Type",
      y     = "Count"
    )
}
```

```{r isar_top_table, eval=!is.null(params$isar_dir) && nrow(isar_switches) > 0}
cols_to_show <- intersect(
  c("gene_name", "gene_id", "isoform_id", "condition_1", "condition_2",
    "IF1", "IF2", "dIF", "isoform_switch_q_value", "gene_switch_q_value", "significant"),
  colnames(isar_switches)
)

if (length(cols_to_show) > 0) {
  display_df <- isar_switches %>% select(all_of(cols_to_show))
  has_sig     <- "significant" %in% colnames(display_df)
  n_sig_table <- if (has_sig) sum(display_df$significant, na.rm = TRUE) else 0

  dt <- datatable(
    display_df,
    caption  = paste0(
      "All tested isoforms, ordered by q-value. ",
      if (has_sig) paste0(n_sig_table, " pass FDR < ", params$fdr_cutoff,
                          " & |dIF| >= ", params$dpsi_cutoff, ".") else ""
    ),
    options  = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    formatRound(columns = names(display_df)[sapply(display_df, is.numeric)], digits = 4)

  if (has_sig) {
    dt <- dt %>%
      formatStyle(
        "significant",
        target          = "row",
        backgroundColor = styleEqual(c(TRUE, FALSE), c("#d4edda", "white"))
      )
  }
  dt
}
```


---

# LeafCutter Results

```{r leafcutter_load, eval=!is.null(params$leafcutter_dir)}
# Standard leafcutter_ds.R output: *_cluster_significance.txt and *_effect_sizes.txt
sig_file <- list.files(params$leafcutter_dir,
                       pattern = "_cluster_significance\\.txt$",
                       full.names = TRUE)
sig_file <- if (length(sig_file) > 0) sig_file[1] else NA_character_

eff_file <- list.files(params$leafcutter_dir,
                       pattern = "_effect_sizes\\.txt$",
                       full.names = TRUE)
eff_file <- if (length(eff_file) > 0) eff_file[1] else NA_character_

if (!is.na(sig_file) && file.exists(sig_file)) {
  leafcutter_clusters <- tryCatch(
    read.delim(sig_file, stringsAsFactors = FALSE),
    error = function(e) data.frame()
  )
  if (nrow(leafcutter_clusters) > 0 && "p.adjust" %in% colnames(leafcutter_clusters)) {
    leafcutter_clusters <- leafcutter_clusters %>%
      mutate(Significant = p.adjust < params$fdr_cutoff & status == "Success")
  }
}

# Sanity check: leafcutter_ds.R writes status = "Success"; flag silently on any
# future vocabulary change so a mismatch cannot pass unnoticed.
if (nrow(leafcutter_clusters) > 0 &&
    sum(leafcutter_clusters$status %in% c("OK", "Success"), na.rm = TRUE) == 0) {
  warning("LeafCutter: no clusters with status OK/Success — possible vocabulary change in leafcutter_ds.R")
}

leafcutter_effects <- data.frame()
if (!is.na(eff_file) && file.exists(eff_file)) {
  leafcutter_effects <- tryCatch(
    read.delim(eff_file, stringsAsFactors = FALSE),
    error = function(e) data.frame()
  )
}
```

```{r leafcutter_summary, eval=!is.null(params$leafcutter_dir) && nrow(leafcutter_clusters) > 0}
n_clusters <- nrow(leafcutter_clusters)
n_ok       <- sum(leafcutter_clusters$status == "Success", na.rm = TRUE)
n_sig      <- sum(leafcutter_clusters$Significant, na.rm = TRUE)

lc_summary_rows <- data.frame(
  Metric = c(
    "Total clusters tested",
    "Clusters passing noise filter (status = Success)",
    sprintf("Significant clusters (FDR < %.2f)", params$fdr_cutoff)
  ),
  Value = c(n_clusters, n_ok, n_sig),
  stringsAsFactors = FALSE
)

if ("genes" %in% colnames(leafcutter_clusters)) {
  sig_genes_lc <- unique(unlist(strsplit(
    leafcutter_clusters$genes[leafcutter_clusters$Significant & !is.na(leafcutter_clusters$genes)],
    ","
  )))
  sig_genes_lc <- trimws(sig_genes_lc[sig_genes_lc != "" & !is.na(sig_genes_lc)])
  lc_summary_rows <- rbind(
    lc_summary_rows,
    data.frame(Metric = "Genes in significant clusters", Value = length(sig_genes_lc),
               stringsAsFactors = FALSE)
  )
}

knitr::kable(lc_summary_rows, caption = "LeafCutter summary")
```

```{r leafcutter_dpsi_dist, eval=!is.null(params$leafcutter_dir) && nrow(leafcutter_effects) > 0 && "deltapsi" %in% colnames(leafcutter_effects), fig.height=5}
leafcutter_effects %>%
  filter(!is.na(deltapsi)) %>%
  ggplot(aes(x = abs(deltapsi))) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  theme_bw() +
  labs(
    title = "Distribution of |ΔΨ| across LeafCutter introns",
    x     = "|ΔΨ|",
    y     = "Count"
  )
```

```{r leafcutter_top_table, eval=!is.null(params$leafcutter_dir) && nrow(leafcutter_clusters) > 0}
cols_to_show <- intersect(
  c("cluster", "genes", "status", "loglr", "df", "p", "p.adjust", "Significant"),
  colnames(leafcutter_clusters)
)

if (length(cols_to_show) > 0) {
  top_clusters <- leafcutter_clusters %>%
    filter(status == "Success") %>%
    arrange(p.adjust) %>%
    select(all_of(cols_to_show)) %>%
    head(100)

  if (nrow(top_clusters) > 0) {
    datatable(
      top_clusters,
      caption  = paste0("Top 100 LeafCutter clusters (status = Success), sorted by FDR. ",
                        n_sig, " pass FDR < ", params$fdr_cutoff, "."),
      options  = list(pageLength = 15, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      formatRound(columns = intersect(c("loglr", "p", "p.adjust"), cols_to_show), digits = 4)
  } else {
    cat("No clusters passed the noise filter.\n")
  }
}
```


---

# Cross-Tool Overlap

```{r overlap_build}
# Build per-tool gene sets of significant AS genes.
# Symbols are normalised (trim/case) and, when a symbol maps to a unique ENSG,
# the ENSG is used as the canonical key so that case/alias mismatches across
# tools do not inflate (or deflate) the apparent overlap.
gene_sets <- list()
active_tool_sig_sets <- list()
active_tool_sig_symbols <- list()

# --- symbol -> ENSG dictionary (unique mappings only) ---
pair_list <- list()
if (nrow(all_rmats) > 0 && all(c("geneSymbol", "GeneID") %in% colnames(all_rmats))) {
  pair_list[["rMATS"]] <- data.frame(sym = as.character(all_rmats$geneSymbol),
                                     gid = as.character(all_rmats$GeneID),
                                     stringsAsFactors = FALSE)
}
if (nrow(majiq_results) > 0 && all(c("gene_name", "gene_id") %in% colnames(majiq_results))) {
  pair_list[["MAJIQ"]] <- data.frame(sym = as.character(majiq_results$gene_name),
                                     gid = as.character(majiq_results$gene_id),
                                     stringsAsFactors = FALSE)
}
if (nrow(isar_switches) > 0 && all(c("gene_name", "gene_id") %in% colnames(isar_switches))) {
  pair_list[["ISAR"]] <- data.frame(sym = as.character(isar_switches$gene_name),
                                    gid = as.character(isar_switches$gene_id),
                                    stringsAsFactors = FALSE)
}
pairs <- if (length(pair_list) > 0) {
  do.call(rbind, pair_list)
} else {
  data.frame(sym = character(), gid = character(), stringsAsFactors = FALSE)
}
pairs$sym <- toupper(trimws(pairs$sym))
pairs$gid <- trimws(as.character(pairs$gid))
pairs <- pairs[!is.na(pairs$sym) & pairs$sym != "" & !is.na(pairs$gid) &
               pairs$gid != "" & grepl("^ENS", pairs$gid), , drop = FALSE]

sym_to_ensg <- list()
ensg_to_sym <- list()
if (nrow(pairs) > 0) {
  n_uniq <- tapply(pairs$gid, pairs$sym, function(x) length(unique(x)))
  pairs <- pairs[pairs$sym %in% names(n_uniq)[n_uniq == 1], , drop = FALSE]
  pairs <- pairs[!duplicated(pairs$sym), , drop = FALSE]
  sym_to_ensg <- as.list(setNames(pairs$gid, pairs$sym))
  ensg_to_sym <- as.list(setNames(pairs$sym, pairs$gid))
}

canonical_key <- function(x) {
  s <- toupper(trimws(as.character(x)))
  vapply(s, function(k) {
    if (nzchar(k) && !is.null(sym_to_ensg[[k]])) sym_to_ensg[[k]] else k
  }, character(1))
}
key_to_display <- function(k) {
  vapply(as.character(k), function(z) {
    if (!is.null(ensg_to_sym[[z]])) ensg_to_sym[[z]] else toupper(trimws(z))
  }, character(1))
}

# rMATS
rmats_sig_genes <- character(0)
rmats_sig_keys <- character(0)
if (nrow(all_rmats) > 0 && "geneSymbol" %in% colnames(all_rmats)) {
  rmats_sig_genes <- unique(all_rmats$geneSymbol[
    all_rmats$Significant & !is.na(all_rmats$geneSymbol) & all_rmats$geneSymbol != ""
  ])
  if (length(rmats_sig_genes) > 0) {
    rmats_sig_keys <- canonical_key(rmats_sig_genes)
    gene_sets[["rMATS"]] <- rmats_sig_keys
    active_tool_sig_symbols[["rMATS"]] <- toupper(trimws(rmats_sig_genes))
  }
}
if (!is.null(params$rmats_dir)) active_tool_sig_sets[["rMATS"]] <- rmats_sig_keys

# MAJIQ
majiq_sig_genes <- character(0)
majiq_sig_keys <- character(0)
if (nrow(majiq_results) > 0 && "gene_name" %in% colnames(majiq_results)) {
  majiq_mask <- majiq_results$Significant & !is.na(majiq_results$gene_name) & trimws(majiq_results$gene_name) != ""
  if ("InformativeGene" %in% colnames(majiq_results)) {
    majiq_mask <- majiq_mask & majiq_results$InformativeGene
  }
  majiq_sig_genes <- unique(majiq_results$gene_name[majiq_mask])
  if (length(majiq_sig_genes) > 0) {
    majiq_sig_keys <- canonical_key(majiq_sig_genes)
    gene_sets[["MAJIQ"]] <- majiq_sig_keys
    active_tool_sig_symbols[["MAJIQ"]] <- toupper(trimws(majiq_sig_genes))
  }
}
if (!is.null(params$majiq_dir)) active_tool_sig_sets[["MAJIQ"]] <- majiq_sig_keys

# ISAR
isar_sig_genes <- character(0)
isar_sig_keys <- character(0)
if (nrow(isar_switches) > 0 && "gene_name" %in% colnames(isar_switches) &&
    "significant" %in% colnames(isar_switches)) {
  isar_sig_genes <- unique(isar_switches$gene_name[
    isar_switches$significant & !is.na(isar_switches$gene_name)
  ])
  if (length(isar_sig_genes) > 0) {
    isar_sig_keys <- canonical_key(isar_sig_genes)
    gene_sets[["ISAR"]] <- isar_sig_keys
    active_tool_sig_symbols[["ISAR"]] <- toupper(trimws(isar_sig_genes))
  }
}
if (!is.null(params$isar_dir)) active_tool_sig_sets[["ISAR"]] <- isar_sig_keys

# LeafCutter (no ENSG) — normalise symbols and map through the dictionary when they match
leafcutter_sig_genes <- character(0)
leafcutter_sig_keys <- character(0)
if (nrow(leafcutter_clusters) > 0 &&
    "genes" %in% colnames(leafcutter_clusters) &&
    "Significant" %in% colnames(leafcutter_clusters)) {
  lc_genes_raw <- leafcutter_clusters$genes[
    leafcutter_clusters$Significant & !is.na(leafcutter_clusters$genes)
  ]
  leafcutter_sig_genes <- unique(trimws(unlist(strsplit(lc_genes_raw, ","))))
  leafcutter_sig_genes <- leafcutter_sig_genes[leafcutter_sig_genes != ""]
  if (length(leafcutter_sig_genes) > 0) {
    leafcutter_sig_keys <- canonical_key(leafcutter_sig_genes)
    gene_sets[["LeafCutter"]] <- leafcutter_sig_keys
    active_tool_sig_symbols[["LeafCutter"]] <- toupper(trimws(leafcutter_sig_genes))
  }
}
if (!is.null(params$leafcutter_dir)) active_tool_sig_sets[["LeafCutter"]] <- leafcutter_sig_keys

# Robust cross-tool core: intersect canonical keys; report display symbols
shared_sig_keys <- if (length(active_tool_sig_sets) >= 2) {
  Reduce(intersect, active_tool_sig_sets)
} else {
  character(0)
}
shared_sig_genes <- if (length(shared_sig_keys) > 0) {
  unique(key_to_display(shared_sig_keys))
} else {
  character(0)
}

# Tools with data analysed but no significant gene set — flagged so a silent
# drop from the UpSet is never mistaken for a missing tool.
empty_tool_sets <- setdiff(names(active_tool_sig_sets), names(gene_sets))
```

```{r overlap_note, results='asis'}
# The UpSet plot below is a gene-level overlap, not an event-level one.
# rMATS reports exon-level events, MAJIQ reports local splice variations
# (LSVs, junction-level), IsoformSwitchAnalyzeR reports whole isoform
# switches, and LeafCutter reports intron clusters. Two tools sharing a
# gene does NOT mean they detected the same event — they may be capturing
# distinct aspects of the same gene's regulation. For event-level
# concordance, downstream matching by genomic coordinates is required.

# Warn when a tool ran but contributed no significant genes to the UpSet.
if (length(empty_tool_sets) > 0) {
  cat(sprintf(
    "<p style='color:#b30000'>WARNING: %d analysed tool(s) contributed no significant genes and are excluded from the UpSet: %s.</p>\n\n",
    length(empty_tool_sets), paste(empty_tool_sets, collapse = ", ")
  ))
}
```

```{r overlap_set_sizes, eval=length(gene_sets) >= 1}
set_sizes <- data.frame(
  Tool      = names(gene_sets),
  Genes_sig = lengths(gene_sets)
)
datatable(
  set_sizes,
  caption  = "Significant AS genes per tool (overlap is at gene level, not event level)",
  options  = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

```{r overlap_upset, eval=length(gene_sets) >= 2, fig.height=5}
upset(
  fromList(gene_sets),
  order.by    = "freq",
  sets.bar.color = "steelblue",
  main.bar.color = "steelblue"
)
```

```{r overlap_summary, eval=length(active_tool_sig_sets) >= 2}
# Genes found significant by all available tools (canonical keys), shown as symbols.
# Count over gene_sets (tools that actually contribute signal), not
# active_tool_sig_sets (tools that ran), so the denominator is not inflated.
all_tools_genes_keys <- Reduce(intersect, gene_sets)
cat(sprintf("Genes significant in all %d tools with signal: **%d**\n\n",
            length(gene_sets), length(all_tools_genes_keys)))

if (length(all_tools_genes_keys) > 0) {
  shared_genes_tbl <- data.frame(
    Gene = sort(key_to_display(all_tools_genes_keys)),
    stringsAsFactors = FALSE
  )
  datatable(
    shared_genes_tbl,
    caption = "Genes significant in all active tools (searchable)",
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE
  )
}
```

```{r overlap_pairwise, eval=length(gene_sets) >= 2}
# Pairwise overlap table
tool_names <- names(gene_sets)
overlap_mat <- matrix(
  nrow = length(tool_names), ncol = length(tool_names),
  dimnames = list(tool_names, tool_names)
)
for (i in tool_names) {
  for (j in tool_names) {
    overlap_mat[i, j] <- length(intersect(gene_sets[[i]], gene_sets[[j]]))
  }
}
kable(as.data.frame(overlap_mat),
      caption = "Pairwise gene overlap between tools (number of shared significant genes)")
```

## Shared-genes ORA (intersection of active tools)

```{r overlap_shared_ora, results='asis'}
if (length(active_tool_sig_sets) < 2) {
  cat("Shared-genes ORA requires at least two active tools.\n")
} else if (length(shared_sig_genes) == 0) {
  cat("No genes available for shared-genes ORA.\n")
} else {
  local_truncate_dt <- function(df, caption) {
    geneid_col <- which(colnames(df) == "geneID")
    dt_options <- list(pageLength = 15, scrollX = TRUE)
    if (length(geneid_col) == 1) {
      dt_options$columnDefs <- list(list(
        targets = geneid_col - 1,
        render = JS(
          "function(data, type, row) {",
          "  if (type !== 'display' || data === null || data === undefined) return data;",
          "  if (data.length <= 80) return data;",
          "  return '<span title=\"' + data + '\">' + data.substr(0, 80) + '... <em>(see more)</em></span>';",
          "}"
        )
      ))
    }
    datatable(df, caption = caption, options = dt_options, rownames = FALSE, escape = FALSE)
  }

  shared_sig_genes <- unique(shared_sig_genes)
  shared_sig_genes <- shared_sig_genes[!is.na(shared_sig_genes) & shared_sig_genes != ""]

  cat(sprintf("Input significant genes: %d\n\n", length(shared_sig_genes)))

  if (length(shared_sig_genes) < 10) {
    cat(sprintf("Need at least 10 genes for shared-genes ORA; found %d.\n", length(shared_sig_genes)))
  } else if (!requireNamespace("clusterProfiler", quietly = TRUE)) {
    cat("clusterProfiler is not available in the report environment.\n")
  } else {
    orgdb_name <- if (identical(params$organism, "mouse")) "org.Mm.eg.db" else "org.Hs.eg.db"
    if (!requireNamespace(orgdb_name, quietly = TRUE)) {
      cat(sprintf("%s is not available in the report environment.\n", orgdb_name))
    } else {
      orgdb_obj <- get(orgdb_name, envir = asNamespace(orgdb_name))

      ego_shared <- tryCatch(
        clusterProfiler::enrichGO(
          gene = shared_sig_genes,
          OrgDb = orgdb_obj,
          keyType = "SYMBOL",
          ont = "BP",
          pAdjustMethod = "BH",
          pvalueCutoff = 0.05,
          qvalueCutoff = 0.2,
          readable = TRUE
        ),
        error = function(e) NULL
      )

      cat("**GO Biological Process**\n\n")
      if (!is.null(ego_shared) && nrow(as.data.frame(ego_shared)) > 0) {
          print(style_ora_dotplot(
            clusterProfiler::dotplot(ego_shared, showCategory = 15),
            "GO BP enrichment (shared genes)"
          ))

        ego_shared_df <- as.data.frame(ego_shared) %>%
          dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
          mutate(across(c(pvalue, p.adjust), ~ signif(.x, 3)))

        print(local_truncate_dt(
          ego_shared_df,
          caption = "GO BP enrichment table - shared genes"
        ))
      } else {
        cat("No GO BP terms available for shared genes.\n\n")
      }

      shared_ids <- tryCatch(
        clusterProfiler::bitr(
          shared_sig_genes,
          fromType = "SYMBOL",
          toType = "ENTREZID",
          OrgDb = orgdb_obj
        ),
        error = function(e) data.frame()
      )
      shared_ids <- unique(shared_ids$ENTREZID)
      shared_ids <- shared_ids[!is.na(shared_ids) & shared_ids != ""]

      cat("**KEGG pathways**\n\n")
      if (length(shared_ids) >= 5) {
        kegg_org <- if (identical(params$organism, "mouse")) "mmu" else "hsa"
        ekegg_shared <- tryCatch(
          clusterProfiler::enrichKEGG(
            gene = shared_ids,
            organism = kegg_org,
            pAdjustMethod = "BH",
            pvalueCutoff = 0.05
          ),
          error = function(e) NULL
        )

        if (!is.null(ekegg_shared) && nrow(as.data.frame(ekegg_shared)) > 0) {
          print(style_ora_dotplot(
            clusterProfiler::dotplot(ekegg_shared, showCategory = 15),
            "KEGG enrichment (shared genes)"
          ))

          ekegg_shared_df <- as.data.frame(ekegg_shared) %>%
            dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
            mutate(across(c(pvalue, p.adjust), ~ signif(.x, 3)))

          print(local_truncate_dt(
            ekegg_shared_df,
            caption = "KEGG enrichment table - shared genes"
          ))
        } else {
          cat("No KEGG pathways available for shared genes.\n\n")
        }
      } else {
        cat("Insufficient SYMBOL-to-ENTREZ mapping for KEGG (need at least 5 genes).\n\n")
      }
    }
  }
}
```

---

# Pathway Enrichment

```{r enrichment_helpers}
get_orgdb_name <- function(organism) {
  if (identical(organism, "mouse")) "org.Mm.eg.db" else "org.Hs.eg.db"
}

truncate_geneid_datatable <- function(df, caption) {
  geneid_col <- which(colnames(df) == "geneID")

  dt_options <- list(pageLength = 15, scrollX = TRUE)
  if (length(geneid_col) == 1) {
    dt_options$columnDefs <- list(list(
      targets = geneid_col - 1,
      render = JS(
        "function(data, type, row) {",
        "  if (type !== 'display' || data === null || data === undefined) return data;",
        "  if (data.length <= 80) return data;",
        "  return '<span title=\"' + data + '\">' + data.substr(0, 80) + '... <em>(see more)</em></span>';",
        "}"
      )
    ))
  }

  datatable(
    df,
    caption = caption,
    options = dt_options,
    rownames = FALSE,
    escape = FALSE
  )
}

run_ora_for_set <- function(genes, set_name, min_genes = 10) {
  genes <- unique(as.character(genes))
  genes <- genes[!is.na(genes) & genes != ""]

  res <- list(
    set_name = set_name,
    n_genes = length(genes),
    message = NULL,
    notes = character(0),
    go = NULL,
    go_df = data.frame(),
    kegg = NULL,
    kegg_df = data.frame()
  )

  if (length(genes) < min_genes) {
    res$message <- sprintf("Need at least %d genes for ORA; found %d.", min_genes, length(genes))
    return(res)
  }

  orgdb_name <- get_orgdb_name(params$organism)
  if (!requireNamespace("clusterProfiler", quietly = TRUE)) {
    res$message <- "clusterProfiler is not available in the report environment."
    return(res)
  }
  if (!requireNamespace(orgdb_name, quietly = TRUE)) {
    res$message <- sprintf("%s is not available in the report environment.", orgdb_name)
    return(res)
  }

  orgdb_obj <- get(orgdb_name, envir = asNamespace(orgdb_name))

  res$go <- tryCatch(
    clusterProfiler::enrichGO(
      gene = genes,
      OrgDb = orgdb_obj,
      keyType = "SYMBOL",
      ont = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2,
      readable = TRUE
    ),
    error = function(e) NULL
  )

  if (!is.null(res$go) && nrow(as.data.frame(res$go)) > 0) {
    res$go_df <- as.data.frame(res$go) %>%
      dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
      mutate(across(c(pvalue, p.adjust), ~ signif(.x, 3)))
  } else {
    res$notes <- c(res$notes, "No GO BP terms passed enrichment thresholds.")
  }

  gene_ids <- tryCatch(
    clusterProfiler::bitr(
      genes,
      fromType = "SYMBOL",
      toType = "ENTREZID",
      OrgDb = orgdb_obj
    ),
    error = function(e) data.frame()
  )
  gene_ids <- unique(gene_ids$ENTREZID)
  gene_ids <- gene_ids[!is.na(gene_ids) & gene_ids != ""]

  if (length(gene_ids) >= 5) {
    kegg_org <- if (identical(params$organism, "mouse")) "mmu" else "hsa"
    res$kegg <- tryCatch(
      clusterProfiler::enrichKEGG(
        gene = gene_ids,
        organism = kegg_org,
        pAdjustMethod = "BH",
        pvalueCutoff = 0.05
      ),
      error = function(e) NULL
    )

    if (!is.null(res$kegg) && nrow(as.data.frame(res$kegg)) > 0) {
      res$kegg_df <- as.data.frame(res$kegg) %>%
        dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
        mutate(across(c(pvalue, p.adjust), ~ signif(.x, 3)))
    } else {
      res$notes <- c(res$notes, "No KEGG pathways passed enrichment thresholds.")
    }
  } else {
    res$notes <- c(res$notes, "Insufficient SYMBOL-to-ENTREZ mapping for KEGG (need at least 5 genes).")
  }

  res
}

render_ora_result <- function(ora_res, heading_level = 3) {
  heading_prefix <- paste(rep("#", heading_level), collapse = "")
  cat(sprintf("%s %s\n\n", heading_prefix, ora_res$set_name))
  cat(sprintf("Input significant genes: %d\n\n", ora_res$n_genes))

  if (!is.null(ora_res$message)) {
    cat(ora_res$message, "\n\n")
    return(invisible(NULL))
  }

  cat("**GO Biological Process**\n\n")
  if (!is.null(ora_res$go) && nrow(ora_res$go_df) > 0) {
    print(style_ora_dotplot(
      clusterProfiler::dotplot(ora_res$go, showCategory = 15),
      paste0("GO BP enrichment (", ora_res$set_name, ")")
    ))
    print(truncate_geneid_datatable(
      ora_res$go_df,
      caption = paste0("GO BP enrichment table — ", ora_res$set_name)
    ))
  } else {
    cat("No GO BP terms available for this gene set.\n\n")
  }

  cat("**KEGG pathways**\n\n")
  if (!is.null(ora_res$kegg) && nrow(ora_res$kegg_df) > 0) {
    print(style_ora_dotplot(
      clusterProfiler::dotplot(ora_res$kegg, showCategory = 15),
      paste0("KEGG enrichment (", ora_res$set_name, ")")
    ))
    print(truncate_geneid_datatable(
      ora_res$kegg_df,
      caption = paste0("KEGG enrichment table — ", ora_res$set_name)
    ))
  } else {
    cat("No KEGG pathways available for this gene set.\n\n")
  }

  if (length(ora_res$notes) > 0) {
    cat(paste0("Notes: ", paste(ora_res$notes, collapse = " ")), "\n\n")
  }
}
```

## ORA by tool

```{r enrichment_by_tool, results='asis'}
if (length(active_tool_sig_sets) == 0) {
  cat("No active tools available for tool-specific ORA.\n")
} else {
  for (tool_name in names(active_tool_sig_sets)) {
    tool_genes <- if (!is.null(active_tool_sig_symbols[[tool_name]])) {
      active_tool_sig_symbols[[tool_name]]
    } else {
      character(0)
    }
    tool_res <- run_ora_for_set(
      tool_genes,
      paste0(tool_name, " significant genes")
    )
    render_ora_result(tool_res, heading_level = 3)
  }
}
```

## ORA global (union of active tools)

```{r enrichment_global, results='asis'}
global_sig_genes <- unique(unlist(active_tool_sig_symbols, use.names = FALSE))
global_res <- run_ora_for_set(global_sig_genes, "Global AS genes (union across active tools)")
render_ora_result(global_res, heading_level = 3)
```

---

# PEGASAS: Pathway–Splicing Correlation

```{r pegasas_check}
run_pegasas_section <- !is.null(params$pegasas_dir)
```

```{r pegasas_intro, eval=run_pegasas_section, results='asis'}
cat("PEGASAS asks a different question than a standard DE pathway enrichment: ",
    "it does not compare patient vs healthy directly. For each sample, it computes ",
    "a single-sample KS enrichment score per pathway (how skewed that gene set is ",
    "in that sample's own expression ranking), then correlates this per-sample score ",
    "against per-sample PSI values, across samples, to find splicing events that ",
    "track with pathway activity. The two plots below show the raw per-sample scores; ",
    "they are **not** a patient-vs-healthy differential ranking. The actual ",
    "splicing–pathway association is the correlation table further down.\n\n")
```

```{r pegasas_load, eval=run_pegasas_section}
# PEGASAS runs pathway enrichment once for the whole cohort and subsets per
# contrast. The per-contrast pegasas directory holds:
#   all_pathways_scores.tsv  — long pathway x sample KS scores (for the plots)
#   correlation_out/<pathway>/<pathway>_global_cor_matrix.txt — ALL events (r, p, significant)

scores_file <- file.path(params$pegasas_dir, "all_pathways_scores.tsv")
corr_base   <- file.path(params$pegasas_dir, "correlation_out")
corr_items  <- if (dir.exists(corr_base)) list.files(corr_base, full.names = TRUE, no.. = TRUE) else character(0)
corr_dirs   <- corr_items[dir.exists(corr_items)]   # only directories (reliable with symlinks)

score_df <- data.frame()
if (file.exists(scores_file)) {
  score_df <- tryCatch(
    read.table(scores_file, sep = "\t", header = TRUE, quote = "",
               stringsAsFactors = FALSE, check.names = FALSE),
    error = function(e) data.frame()
  )
}
if (nrow(score_df) > 0) {
  # Safe -log10(p): guard against p == 0 (Inf). KS p-values are always numeric.
  score_df$neglogp <- -log10(pmax(as.numeric(score_df$p_value), .Machine$double.xmin))
}

# Per-pathway correlation tables (ALL events, including non-significant, so the
# section is never empty and it is clear which pathways were tested).
cor_list <- list()
for (cdir in corr_dirs) {
  sig_name <- basename(cdir)
  gfile <- file.path(cdir, paste0(sig_name, "_global_cor_matrix.txt"))
  if (!file.exists(gfile)) next
  d <- tryCatch(
    read.table(gfile, sep = "\t", header = TRUE, quote = "", fill = TRUE,
               stringsAsFactors = FALSE, check.names = FALSE),
    error = function(e) NULL
  )
  if (is.null(d) || nrow(d) == 0) next
  d$pathway <- sig_name
  d$significant <- as.logical(d$significant)
  cor_list[[sig_name]] <- d
}

# BH over the permutation p-values of each pathway. p is NA where the
# permutation was not run (|r| <= 0.3); p.adjust() would propagate those NAs,
# so adjust only over the tested rows and keep NA for the untested ones.
cor_list <- lapply(cor_list, function(d) {
  d$p_adj <- NA_real_
  tested  <- !is.na(d$p_value)
  if (any(tested)) {
    d$p_adj[tested] <- p.adjust(d$p_value[tested], method = "BH")
  }
  d
})

# Keep the most informative rows per pathway (top by |r|) for an HTML that stays
# responsive. The complete, unfiltered table is on disk in correlation_out/.
TOP_N <- 100
cor_display <-
  if (length(cor_list) > 0) do.call(rbind, lapply(cor_list, function(d) {
    d <- d[order(-abs(d$pearson_r)), , drop = FALSE]
    if (nrow(d) > TOP_N) d <- d[seq_len(TOP_N), , drop = FALSE]
    d
  })) else data.frame()

if (nrow(cor_display) > 0) {
  cor_display$event <-
    paste0(cor_display$GeneName, ":",
           cor_display$chr, ":", cor_display$strand, ":",
           cor_display$exonStart, "-", cor_display$exonEnd)
}

# Compact overview: events tested / significant / strongest correlation per pathway.
pathway_summary <- if (length(cor_list) > 0) {
  do.call(rbind, lapply(names(cor_list), function(p) {
    d <- cor_list[[p]]
    ar <- abs(d$pearson_r)
    pv <- d$p_value
    pa <- d$p_adj
    data.frame(
      pathway      = p,
      n_events     = nrow(d),
      n_significant = sum(d$significant, na.rm = TRUE),
      max_abs_r    = if (any(is.finite(ar))) max(ar, na.rm = TRUE) else NA_real_,
      min_p        = if (any(is.finite(pv))) min(pv, na.rm = TRUE) else NA_real_,
      # BH-adjusted p: only the tested events are counted, so NAs in p_value
      # never inflate (or deflate) the per-pathway statistics.
      min_p_adj    = if (any(is.finite(pa))) min(pa, na.rm = TRUE) else NA_real_,
      n_sig_adj    = if (any(!is.na(pa))) sum(pa < 0.05, na.rm = TRUE) else 0L,
      stringsAsFactors = FALSE, check.names = FALSE
    )
  }))
} else data.frame()
```

```{r pegasas_score_plotA_caption, eval=run_pegasas_section && nrow(score_df) > 0, results='asis'}
cat("**Plot A — per-sample pathway activity.** Each dot is one pathway in one sample: ",
    "size = KS enrichment score, colour = -log10(p) of that single-sample test. ",
    "Samples are grouped by condition only for display; this plot does not test for ",
    "a difference between groups.\n\n")
```

```{r pegasas_score_plotA, eval=run_pegasas_section && nrow(score_df) > 0, fig.height=7}
# Heatmap of points (bubble heatmap): pathways x samples, sized by KS score,
# coloured by -log10(p). Samples are faceted by group so the activity pattern
# can be compared across the two contrast groups at a glance.
plot_df <- score_df
plot_df$group  <- factor(plot_df$group, levels = unique(plot_df$group))
sample_order <- unique(plot_df[order(plot_df$group, plot_df$sample), "sample"])
plot_df$sample <- factor(plot_df$sample, levels = sample_order)
path_levels <- names(sort(tapply(plot_df$KS_score, plot_df$pathway, mean)))
plot_df$pathway <- factor(plot_df$pathway, levels = path_levels)

ggplot(plot_df, aes(x = sample, y = pathway,
                    size = KS_score, color = neglogp)) +
  geom_point() +
  scale_size_continuous(range = c(1, 6), name = "KS score") +
  scale_color_gradient(low = "#deebf7", high = "#08306b", name = "-log10(p)") +
  facet_grid(~ group, scales = "free_x", space = "free_x", switch = "x") +
  labs(title = "Pathway KS activity per sample",
       x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        panel.grid = element_blank(),
        strip.background = element_rect(fill = "grey90", color = NA))
```

```{r pegasas_score_plotB_caption, eval=run_pegasas_section && nrow(score_df) > 0, results='asis'}
cat("**Plot B — cohort-average pathway ranking.** Pathways ordered by their mean KS ",
    "score across ALL samples (both groups pooled). This ranks how consistently ",
    "active a pathway is across the cohort — it is **not** a patient-vs-healthy ",
    "differential GSEA and does not rank pathways by group difference.\n\n")
```

```{r pegasas_score_plotB, eval=run_pegasas_section && nrow(score_df) > 0, fig.height=7}
# Ranked dot-plot (GSEA-style): one axis, pathways ordered by mean KS score;
# colour = mean KS, size = mean -log10(p). Compact summary of pathway activity.
rank_df <- aggregate(cbind(KS_score, neglogp) ~ pathway,
                     data = score_df, FUN = mean, na.rm = TRUE)
rank_df <- rank_df[order(rank_df$KS_score), , drop = FALSE]
rank_df$pathway <- factor(rank_df$pathway, levels = rank_df$pathway)

ggplot(rank_df, aes(x = KS_score, y = pathway, color = KS_score, size = neglogp)) +
  geom_point() +
  scale_color_gradient(low = "#deebf7", high = "#08306b", name = "mean KS") +
  scale_size_continuous(range = c(2, 7), name = "mean -log10(p)") +
  labs(title = "Pathway activity ranking (mean KS score)",
       x = "Mean KS enrichment score", y = NULL) +
  theme_bw(base_size = 9) +
  theme(panel.grid.major.y = element_line(color = "grey90"))
```

```{r pegasas_no_scores, eval=run_pegasas_section && nrow(score_df) == 0}
cat("No PEGASAS pathway scores available for this contrast.\n")
```

```{r pegasas_cor_header, eval=run_pegasas_section && nrow(cor_display) > 0, results='asis'}
cat("## Pathway–Correlated Splicing Events\n\n")
cat("All tested events are shown (including non-significant ones), ranked by |Pearson r|.")
cat(sprintf(" For responsiveness each pathway is capped at the top %d events; the complete, unfiltered table is in `correlation_out/<pathway>/<pathway>_global_cor_matrix.txt`.\n\n", TOP_N))
cat("**Two distinct p-values are reported in this section, do not confuse them:**\n\n")
cat("* The **heatmap p-value** is the KS enrichment of the pathway gene set per sample (can reach very small values because of the gene-set size), not a correlation test. It belongs to the pathway activity plots above.\n\n")
cat("* The **correlation p-value** (`p_value` in the tables below) is the permutation p of each event's Pearson r against the pathway score, computed only for |r| > 0.3 (otherwise it is NA). With 1000 permutations p == 0 means 'more extreme than all permutations'; `p_adj` is its Benjamini–Hochberg correction across all events of that pathway.\n\n")
cat("Significance: |Pearson r| > 0.3 AND permutation p == 0 within the resolution of the test (p < 1/N_perms). This means the observed correlation is more extreme than ALL permutations.\n\n")
```

```{r pegasas_cor_table, eval=run_pegasas_section && nrow(cor_display) > 0}
datatable(
  cor_display[, c("pathway", "event", "GeneName", "chr", "strand",
                  "exonStart", "exonEnd", "pearson_r", "p_value", "p_adj", "significant")],
  caption = "Splicing events correlated with pathway activity (all events, top by |r|)",
  rownames = FALSE, filter = "top",
  options = list(pageLength = 15, scrollX = TRUE)
) |>
  formatRound(c("pearson_r", "p_value", "p_adj"), digits = 4) |>
  formatStyle("significant", target = "row",
              backgroundColor = styleEqual(c(TRUE, FALSE), c("#d9f2d9", "#ffffff")))
```

```{r pegasas_summary_header, eval=run_pegasas_section && nrow(pathway_summary) > 0, results='asis'}
cat("### Pathway summary\n\n")
cat("Candidate splicing events are restricted to rMATS-significant events ",
    "(FDR < 0.05, |ΔΨ| ≥ 0.1). This analysis is exploratory and ",
    "hypothesis-generating: it highlights candidate pathway–splicing ",
    "couplings for follow-up, not a primary statistical claim.\n\n")
```

```{r pegasas_summary_table, eval=run_pegasas_section && nrow(pathway_summary) > 0}
datatable(
  pathway_summary,
  caption = "Events tested, significant, and strongest correlation per pathway (BH-adjusted)",
  rownames = FALSE, options = list(pageLength = 15, scrollX = TRUE)
) |>
  formatRound(c("max_abs_r", "min_p", "min_p_adj"), digits = 4)
```

```{r pegasas_no_cor, eval=run_pegasas_section && nrow(cor_display) == 0, results='asis'}
cat("No PEGASAS correlation results available for this contrast.\n")
```

---



# DE + AS Integration

```{r de_load}
# Differential expression results. Gated by run_de_as and de_results.
# de_results should point to a DGE output directory structured as:
#   de_results/<DGE_comparison>/<DGE_comparison>_DESeq2_results.csv
# Pipeline comparison IDs (e.g. NB4_Scramble_vs_NB4_DDX41_sh1) are
# auto-matched to DGE directory names (e.g. NB4_DDX41_sh1_vs_Scramble)
# by reversing group order and stripping the shared cell-type prefix.
# Expected CSV columns: gene_name (HUGO), log2FoldChange, padj.
# gene_name is used as gene_id to cross-reference with AS gene symbols.
run_de_section <- isTRUE(params$run_de_as) &&
                  !is.null(params$de_results) && nzchar(params$de_results) &&
                  dir.exists(params$de_results)
de_file     <- NULL
de_results  <- data.frame()
de_sig_genes <- character(0)

if (run_de_section) {
  # Build reverse DGE name from pipeline comparison_id.
  # e.g. NB4_Scramble_vs_NB4_DDX41_sh1 → NB4_DDX41_sh1_vs_Scramble
  parts   <- strsplit(params$comparison_id, "_vs_")[[1L]]
  prefix  <- sub("_.*$", "", parts[1L])               # cell-type prefix (NB4, MSCline, MNC, MSC)
  g1_clean <- sub(paste0("^", prefix, "_"), "", parts[1L])
  g2_clean <- sub(paste0("^", prefix, "_"), "", parts[2L])
  dge_name <- paste0(prefix, "_", g2_clean, "_vs_", g1_clean)

  dge_subdirs <- list.dirs(params$de_results, recursive = FALSE, full.names = TRUE)
  # Try: (1) exact DGE name, (2) reversed DGE name, (3) any dir containing both
  # group names
  for (d in dge_subdirs) {
    dn <- basename(d)
    if (dn == dge_name || dn == params$comparison_id ||
        (grepl(g1_clean, dn, fixed = TRUE) && grepl(g2_clean, dn, fixed = TRUE))) {
      csvs <- list.files(d, pattern = "\\.csv$", full.names = TRUE, ignore.case = TRUE)
      if (length(csvs) > 0L) {
        de_file <- csvs[1L]
        break
      }
    }
  }

  if (!is.null(de_file)) {
    de_results <- tryCatch(
      read.csv(de_file, stringsAsFactors = FALSE),
      error = function(e) {
        message("Could not load DE results: ", conditionMessage(e))
        data.frame()
      }
    )
  }
}

# Normalise column names: DESeq2 outputs gene_name, log2FoldChange; the report
# expects gene_id, log2FC, padj.
if (run_de_section && nrow(de_results) > 0L) {
  n <- colnames(de_results)
  # Prefer HUGO gene names for cross-referencing with AS gene symbols.
  # DGE CSVs often have both gene_id (ENSG) and gene_name (HUGO).
  # Drop ENSG gene_id first, then rename gene_name → gene_id to avoid dups.
  if ("gene_name" %in% n && "gene_id" %in% n) {
    de_results[["gene_id"]] <- NULL   # drop ENSG column
  }
  if ("gene_name" %in% n) colnames(de_results)[colnames(de_results) == "gene_name"] <- "gene_id"
  if ("log2FoldChange" %in% n && !"log2FC" %in% n) colnames(de_results)[colnames(de_results) == "log2FoldChange"] <- "log2FC"

  # Validate required columns
  required_cols <- c("gene_id", "log2FC", "padj")
  missing_cols  <- setdiff(required_cols, colnames(de_results))
  if (length(missing_cols) > 0L) {
    warning("DE results file is missing columns: ", paste(missing_cols, collapse = ", "),
            ". DE+AS section will be skipped.")
    run_de_section <- FALSE
    de_results     <- data.frame()
  }
}

if (run_de_section && nrow(de_results) > 0) {
  de_results <- de_results %>%
    filter(!is.na(padj) & !is.na(log2FC)) %>%
    mutate(
      DE_significant = padj < params$fdr_cutoff & abs(log2FC) >= 1,
      neg_log10_padj = -log10(pmax(padj, 1e-300))
    )
  de_sig_genes <- unique(de_results$gene_id[de_results$DE_significant])
}
```

```{r de_no_results, eval=run_de_section && nrow(de_results) == 0, results='asis'}
cat("No matching DE (DESeq2) results were found in `de_results` for comparison `",
    params$comparison_id, "` — DE + AS integration is skipped for this contrast. ",
    "Generate a DGE subdirectory matching this comparison's groups if you need this section.\n\n", sep = "")
```

```{r de_as_overlap, eval=run_de_section && nrow(de_results) > 0}
# Pool AS significant genes from all active tools.
# Reuse the gene sets built in the Cross-Tool Overlap section.
as_sig_genes <- unique(c(rmats_sig_genes, majiq_sig_genes, isar_sig_genes, leafcutter_sig_genes))

as_tool_sets <- list(
  rMATS       = rmats_sig_genes,
  MAJIQ       = majiq_sig_genes,
  ISAR        = isar_sig_genes,
  LeafCutter  = leafcutter_sig_genes
)
# Keep only tools that contribute genes.
active_as_tools <- names(as_tool_sets)[lengths(as_tool_sets) > 0]

# Tag each gene in the DE table by its AS status and per-tool membership.
de_results <- de_results %>%
  mutate(
    AS_hit = gene_id %in% as_sig_genes,
    hit_class = case_when(
      DE_significant & AS_hit  ~ "DE + AS",
      DE_significant & !AS_hit ~ "DE only",
      !DE_significant & AS_hit ~ "AS only",
      TRUE                     ~ "Neither"
    )
  )

# Per-tool AS hit flag (gene detected as significant by each tool).
for (tl in active_as_tools) {
  de_results[[paste0("AS_", tl)]] <- de_results$gene_id %in% as_tool_sets[[tl]]
}

dual_hits <- de_results %>%
  filter(hit_class == "DE + AS") %>%
  arrange(padj)

# Ensures the downstream chunks that reference dual_hits_ann always see it,
# even when there are no DE+AS hits (it is populated by the de_as_table chunk).
dual_hits_ann <- data.frame()

cat(sprintf(
  "DE significant (|log2FC| ≥ 1, padj < %.2f): **%d genes**  \n",
  params$fdr_cutoff, length(de_sig_genes)
))
cat(sprintf(
  "AS significant (any tool, FDR/prob cutoff): **%d genes**  \n",
  length(as_sig_genes)
))
cat(sprintf(
  "Dual hits (DE + AS): **%d genes**\n\n",
  nrow(dual_hits)
))

# Reusable DE volcano plot, colouring AS hits by one tool membership column.
# as_col: column name in de_results flagging the AS hit for the chosen tool.
make_de_volcano <- function(de_df, as_col, title, subtitle) {
  set.seed(42)
  colour_map <- c(
    "DE + AS"  = "firebrick",
    "DE only"  = "steelblue",
    "AS only"  = "darkorange",
    "Neither"  = "gray80"
  )
  size_map   <- c("DE + AS" = 2.5, "DE only" = 1.5, "AS only" = 2, "Neither" = 0.6)
  alpha_map  <- c("DE + AS" = 0.9, "DE only" = 0.6, "AS only" = 0.7, "Neither" = 0.3)

  df <- de_df
  df$tool_hit <- df[[as_col]] %in% TRUE
  df$hh <- ifelse(df$DE_significant & df$tool_hit, "DE + AS",
                  ifelse(df$DE_significant & !df$tool_hit, "DE only",
                         ifelse(!df$DE_significant & df$tool_hit, "AS only", "Neither")))
  df$hh <- factor(df$hh, levels = names(colour_map))

  plot_df <- bind_rows(
    df %>% filter(hh != "Neither"),
    df %>% filter(hh == "Neither") %>% slice_sample(n = min(3000, sum(df$hh == "Neither")))
  )
  label_df <- df %>% filter(hh == "DE + AS") %>%
    slice_max(neg_log10_padj, n = 20)

  ggplot(plot_df, aes(x = log2FC, y = neg_log10_padj,
                      colour = hh, size = hh, alpha = hh)) +
    geom_point() +
    geom_text_repel(
      data = label_df, aes(label = gene_id),
      colour = "black", size = 2.8, max.overlaps = 15, segment.size = 0.3
    ) +
    scale_colour_manual(values = colour_map, name = NULL) +
    scale_size_manual(values = size_map, name = NULL) +
    scale_alpha_manual(values = alpha_map, name = NULL) +
    geom_hline(yintercept = -log10(params$fdr_cutoff),
               linetype = "dashed", colour = "gray40", linewidth = 0.4) +
    geom_vline(xintercept = c(-1, 1),
               linetype = "dashed", colour = "gray40", linewidth = 0.4) +
    theme_bw(base_size = 12) +
    labs(title = title, subtitle = subtitle,
         x = "log2 Fold Change", y = "-log10(adj. p-value)") +
    theme(legend.position = "bottom")
}
```

```{r de_volcano, eval=run_de_section && nrow(de_results) > 0, fig.height=7, fig.width=10}
# Volcano plot with AS-hit genes highlighted.
# Points are down-sampled for "Neither" to keep the plot readable.
set.seed(42)
de_plot <- bind_rows(
  de_results %>% filter(hit_class != "Neither"),
  de_results %>% filter(hit_class == "Neither") %>% slice_sample(n = min(3000, sum(de_results$hit_class == "Neither")))
)

label_df <- de_results %>%
  filter(hit_class == "DE + AS") %>%
  slice_max(neg_log10_padj, n = 20)

colour_map <- c(
  "DE + AS"  = "firebrick",
  "DE only"  = "steelblue",
  "AS only"  = "darkorange",
  "Neither"  = "gray80"
)
size_map   <- c("DE + AS" = 2.5, "DE only" = 1.5, "AS only" = 2, "Neither" = 0.6)
alpha_map  <- c("DE + AS" = 0.9, "DE only" = 0.6, "AS only" = 0.7, "Neither" = 0.3)

ggplot(de_plot, aes(x = log2FC, y = neg_log10_padj,
                    colour = hit_class, size = hit_class, alpha = hit_class)) +
  geom_point() +
  geom_text_repel(
    data        = label_df,
    aes(label   = gene_id),
    colour      = "black",
    size        = 2.8,
    max.overlaps = 15,
    segment.size = 0.3
  ) +
  scale_colour_manual(values = colour_map, name = NULL) +
  scale_size_manual(values = size_map,     name = NULL) +
  scale_alpha_manual(values = alpha_map,   name = NULL) +
  geom_hline(yintercept = -log10(params$fdr_cutoff),
             linetype = "dashed", colour = "gray40", linewidth = 0.4) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", colour = "gray40", linewidth = 0.4) +
  theme_bw(base_size = 12) +
  labs(
    title    = "Differential Expression — AS genes highlighted",
    subtitle = sprintf("Red = significant in both DE and AS (n = %d)", nrow(dual_hits)),
    x        = "log2 Fold Change",
    y        = "-log10(adj. p-value)"
  ) +
  theme(legend.position = "bottom")
```

```{r de_volcano_by_tool_ui, eval=run_de_section && nrow(de_results) > 0 && length(active_as_tools) > 0, fig.show='hide', results='asis'}
# Per-tool DE volcanoes in a tabbed browser. Each tab is a PNG generated from
# the same DE data, highlighting only the genes called significant by that tool,
# so the "DE + AS" signal can be inspected per AS tool without re-running anything.
vol_by_tool <- list()
vol_by_tool[["All tools"]] <- make_de_volcano(
  de_results, "AS_hit",
  "Differential Expression - AS genes highlighted (all tools)",
  sprintf("Red = significant in both DE and AS (n = %d)", nrow(dual_hits))
)

for (tl in active_as_tools) {
  n_dual_t <- sum(de_results$DE_significant & de_results[[paste0("AS_", tl)]], na.rm = TRUE)
  vol_by_tool[[tl]] <- make_de_volcano(
    de_results, paste0("AS_", tl),
    paste0("Differential Expression - AS genes highlighted (", tl, ")"),
    sprintf("Red = significant in both DE and AS (%s), n = %d", tl, n_dual_t)
  )
}
```

```{r de_volcano_by_tool, eval=run_de_section && nrow(de_results) > 0 && length(active_as_tools) > 0, results='asis'}
# Render each volcano to a self-contained PNG and drive visibility with simple tabs.
render_vol_png <- function(gg, width = 9, height = 6, dpi = 110) {
  tmpf <- tempfile(fileext = ".png")
  ggsave(tmpf, plot = gg, width = width, height = height, dpi = dpi, bg = "white")
  sz <- file.info(tmpf)$size
  raw <- readBin(tmpf, what = "raw", n = sz)
  paste0("data:image/png;base64,", jsonlite::base64_enc(raw))
}

widget_id <- paste0("de_volcano_browser_", gsub("[^A-Za-z0-9]+", "_", params$comparison_id))
tool_names <- names(vol_by_tool)

tab_html <- paste0(
  vapply(tool_names, function(tm) {
    sprintf('<button class="%s_tab" data-tool="%s" style="margin:2px 6px 2px 0; padding:4px 12px; cursor:pointer; border:1px solid #ccc; border-radius:4px; background:#f7f7f7;">%s</button>',
            widget_id, tm, tm)
  }, character(1)),
  collapse = ""
)

cat(
  paste0(
    '<div id="', widget_id, '" style="border:1px solid #d9d9d9; border-radius:8px; padding:12px; margin:12px 0 20px 0;">',
    '<div style="font-weight:600; margin-bottom:8px;">DE volcano by AS tool</div>',
    '<div style="margin-bottom:10px;">', tab_html, '</div>'
  )
)

for (tm in tool_names) {
  cat(sprintf(
    '<div class="%s_panel" data-tool="%s" style="display:%s;">',
    widget_id, tm, if (tm == tool_names[1]) "block" else "none"
  ))
  cat(sprintf('<img src="%s" style="max-width:100%%; display:block; margin:0 auto;" alt="%s volcano">',
              render_vol_png(vol_by_tool[[tm]]), tm))
  cat('</div>')
}
cat('</div>')

script <- paste0(
  '<script>(function(){',
  'const w = document.getElementById("', widget_id, '");',
  'const tabs = w.getElementsByClassName("', widget_id, '_tab");',
  'const panels = w.getElementsByClassName("', widget_id, '_panel");',
  'const activate = (tool) => {',
  '  for (const t of tabs) t.style.background = (t.dataset.tool === tool) ? "#337ab7" : "#f7f7f7";',
  '  for (const p of panels) p.style.display = (p.dataset.tool === tool) ? "block" : "none";',
  '};',
  'for (const t of tabs) t.addEventListener("click", () => activate(t.dataset.tool));',
  '})();</script>'
)
cat(script)
```

```{r de_as_table, eval=run_de_section && nrow(de_results) > 0 && nrow(dual_hits) > 0}
# One row per DE+AS gene per AS tool that calls it, so the source tool can be
# filtered/sorted. Per-tool AS metrics summarise the best event in that gene.
tool_rows <- lapply(active_as_tools, function(tl) {
  base <- de_results %>%
    filter(DE_significant & .data[[paste0("AS_", tl)]]) %>%
    select(gene_id, log2FC, padj) %>%
    mutate(Tool = tl, AS_metric = NA_real_)

  if (tl == "rMATS" && nrow(all_rmats) > 0 && "geneSymbol" %in% colnames(all_rmats)) {
    g <- all_rmats %>%
      filter(Significant, geneSymbol %in% base$gene_id) %>%
      group_by(geneSymbol) %>% slice_min(FDR, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(geneSymbol, AS_metric = priority_score)
    base <- base %>% left_join(g, by = c("gene_id" = "geneSymbol"))
  } else if (tl == "MAJIQ" && nrow(majiq_results) > 0 && "gene_name" %in% colnames(majiq_results)) {
    g <- majiq_results %>%
      filter(Significant, gene_name %in% base$gene_id) %>%
      group_by(gene_name) %>% slice_max(probability_changing, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(gene_name, AS_metric = dpsi_mean)
    base <- base %>% left_join(g, by = c("gene_id" = "gene_name"))
  } else if (tl == "LeafCutter" && nrow(leafcutter_clusters) > 0 && "genes" %in% colnames(leafcutter_clusters)) {
    lc_lst <- leafcutter_clusters %>%
      filter(Significant, !is.na(genes)) %>%
      mutate(genes = strsplit(as.character(genes), ",")) %>%
      tidyr::unnest(genes) %>%
      mutate(genes = trimws(genes)) %>%
      filter(genes %in% base$gene_id)
    if (nrow(lc_lst) > 0) {
      g <- lc_lst %>% group_by(genes) %>% slice_min(p.adjust, n = 1, with_ties = FALSE) %>%
        ungroup() %>% select(genes, AS_metric = p.adjust)
      base <- base %>% left_join(g, by = c("gene_id" = "genes"))
    }
  } else if (tl == "ISAR" && nrow(isar_switches) > 0 && "gene_name" %in% colnames(isar_switches)) {
    g <- isar_switches %>%
      filter(significant %in% TRUE, gene_name %in% base$gene_id) %>%
      mutate(AS_metric = abs(dIF)) %>%
      group_by(gene_name) %>% slice_min(isoform_switch_q_value, n = 1, with_ties = FALSE) %>%
      ungroup() %>% select(gene_name, AS_metric)
    base <- base %>% left_join(g, by = c("gene_id" = "gene_name"))
  }
  base
})

dual_hits_ann <- if (length(tool_rows) > 0) {
  bind_rows(tool_rows) %>% distinct(gene_id, Tool, .keep_all = TRUE)
} else {
  data.frame()
}

if (nrow(dual_hits_ann) > 0) {
  show_cols <- intersect(
    c("gene_id", "Tool", "log2FC", "padj", "AS_metric"),
    colnames(dual_hits_ann)
  )
  datatable(
    dual_hits_ann[, show_cols],
    caption  = sprintf("DE + AS genes per source tool (n = %d gene-tool pairs, %d unique genes)",
                       nrow(dual_hits_ann),
                       length(unique(dual_hits_ann$gene_id))),
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 20, scrollX = TRUE)
  ) %>%
    formatRound(columns = intersect(c("log2FC", "padj", "AS_metric"), show_cols), digits = 4)
}
```

```{r de_as_barplot, eval=run_de_section && nrow(de_results) > 0 && nrow(dual_hits_ann) > 0, fig.height=5}
# Distribution of DE+AS genes across the AS tools.
dual_hits_ann %>%
  filter(!is.na(Tool)) %>%
  count(Tool) %>%
  ggplot(aes(x = reorder(Tool, n), y = n, fill = Tool)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_bw(base_size = 12) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "DE+AS genes by source tool",
    x     = NULL,
    y     = "Number of genes"
  )
```

```{r de_as_skip_msg, eval=!run_de_section}
cat("DE+AS integration is disabled.\n")
cat("Enable with two lines in params.yaml:\n")
cat("- `run_de_as: true`\n")
cat("- `de_results: /path/to/dge/`  (directory with subdirs per DGE comparison, each containing `*DESeq2_results.csv`)\n")
```

---

# Data Exports

Paths below are relative to the root of the shipped `deliverables/` folder (the
folder that contains `run_info/` and `contrasts/`), not to this HTML file. This
report and every table it lists ship together; `raw/` is not part of the shipment.

```{r data_exports, eval=!is.null(params$publish_deliverables) && params$publish_deliverables, results='asis'}
# Inventory of the shippable deliverable tables for this comparison. Paths are
# relative to the deliverables/ root (this folder), exactly as recorded in
# contrast_manifest.yaml.
comp <- params$comparison_id
base <- sprintf("contrasts/%s", comp)

export_rows <- data.frame(
  deliverable = character(), description = character(), rel_path = character(),
  tool = character(), status = character(), note = character(),
  stringsAsFactors = FALSE
)

add_export <- function(deliverable, description, rel_path, tool = '', status = '', note = '') {
  export_rows <<- rbind(
    export_rows,
    data.frame(deliverable, description, rel_path, tool, status, note,
               stringsAsFactors = FALSE)
  )
}

active_tool_dirs <- list(
  rMATS = params$rmats_dir, MAJIQ = params$majiq_dir, ISAR = params$isar_dir,
  LeafCutter = params$leafcutter_dir, PEGASAS = params$pegasas_dir
)
tool_key <- list(rMATS = 'rmats', MAJIQ = 'majiq', ISAR = 'isar',
                 LeafCutter = 'leafcutter', PEGASAS = 'pegasas')

for (tool_name in names(active_tool_dirs)) {
  if (is.null(active_tool_dirs[[tool_name]])) next
  key <- tool_key[[tool_name]]
  add_export(
    sprintf("%s master", tool_name),
    "All analysed features with standard columns + native metrics",
    sprintf("%s/tables/%s.%s.master.tsv", base, comp, key),
    tool_name, tool_issue_status(key), tool_issue_note(key)
  )
  add_export(
    sprintf("%s significant", tool_name),
    "Features passing the significance rule used by this report",
    sprintf("%s/tables/%s.%s.significant.tsv", base, comp, key),
    tool_name, tool_issue_status(key), tool_issue_note(key)
  )
  add_export(
    sprintf("%s summary", tool_name),
    "Per-feature-type counts and effect-size distributions",
    sprintf("%s/tables/%s.%s.summary.tsv", base, comp, key),
    tool_name, tool_issue_status(key), tool_issue_note(key)
  )
}

if (!is.null(params$rmats_dir) || any(!is.null(active_tool_dirs[2:4]))) {
  add_export(
    "Cross-tool gene master",
    "Gene x tool union of significant features across AS tools",
    sprintf("%s/tables/%s.cross_tool.master.tsv", base, comp),
    "cross_tool", tool_issue_status('rmats'), ""
  )
  add_export(
    "Cross-tool gene summary",
    "One row per gene with n_tools_significant and the calling tools",
    sprintf("%s/tables/%s.cross_tool.gene_summary.tsv", base, comp),
    "cross_tool", tool_issue_status('rmats'), ""
  )
}
if (!is.null(params$sashimi_dir)) {
  add_export(
    "Sashimi index",
    "Navigable PDF index (plot browsing + publication figures)",
    sprintf("%s/plots/sashimi/sashimi_index.tsv", base),
    "sashimi", tool_issue_status('sashimi'), tool_issue_note('sashimi')
  )
}
add_export(
  "Report",
  "This HTML report (self-contained)",
  sprintf("%s/%s_splicing_report.html", base, comp), "report"
)
add_export(
  "Contrast manifest",
  "YAML record of every deliverable of this comparison (paths relative to deliverables/)",
  sprintf("%s/contrast_manifest.yaml", base), "manifest"
)
add_export(
  "Run manifest",
  "Global run metadata: params, tools, known issues",
  "run_info/run_manifest.yaml", "manifest"
)
add_export(
  "Sample index",
  "Sample x comparison x group table (no cluster paths)",
  "run_info/sample_index.tsv", "manifest"
)
add_export(
  "Software versions",
  "Consolidated tool versions for the whole run",
  "run_info/software_versions.yml", "manifest"
)
add_export(
  "QA report",
  "Results-contract validation output (errors + warnings)",
  "run_info/qa_report.txt", "manifest"
)

show_cols <- c("deliverable", "description", "rel_path", "tool", "status", "note")
display <- export_rows
display$note[display$status == 'ok' | display$note == ''] <- ""

kable(display[, show_cols], caption = sprintf(
  "Shippable deliverables for %s (contract v%s, publish level %s)",
  comp, params$results_contract_version, params$publish_level
))

warn_rows <- display[display$status != 'ok', ]
if (nrow(warn_rows) > 0) {
  cat("<br>**Reliability caveats applying to the tables above:**\n\n")
  for (i in seq_len(nrow(warn_rows))) {
    cat(sprintf("- **%s** (%s): %s\n", warn_rows$tool[i], warn_rows$status[i],
                warn_rows$note[i]))
  }
  cat("\n")
}
```

```{r data_exports_disabled, eval=is.null(params$publish_deliverables) || !params$publish_deliverables, results='asis'}
cat("Deliverables export is disabled (`publish_deliverables: false` in the run params).\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
